# Aggregate Groups {#aggroups}

```{r setup,  echo = F, message=F}

#Load packages
library(knitr)
library(rmarkdown)

```


**Description**: Mappings of species into aggregate group categories for different analyses

**Indicator category**: Synthesis of published information 

**Contributor(s)**: Geret DePiper, Sarah Gaichas, Sean Hardison, Sean Lucey
  
**Data steward**: Sean Lucey <Sean.Lucey@noaa.gov>
  
**Point of contact**: Sean Lucey <Sean.Lucey@noaa.gov>
  
**Public availability statement**: Source data is available to the public


## Methods


### Data source(s) (All types)
Please provide a text description of data sources. 



### Data extraction 
Text overview description of extraction methods. What information was extracted and how was it aggregated? Can point to other indicator extraction methods if the same.



Write SQL query here
```{sql, eval = F, echo = T}
SELECT * FROM...
```

R code used in extraction process.
```{r r_extract, echo = T, eval = F}
# Insert R code used for extraction here
```

### Data analysis
Text description of analysis methods, similar in structure and detail to a peer-reviewed paper methods section.






What packages or libraries did you use in your work flow?
```{r packages, echo = T}
sessionInfo(package = NULL)
#Insert packages used here

#Use this to output a detailed list of the package information
#current.session <- sessionInfo(package = NULL)
#current.session$otherPkgs
```


Include accompanying R code, pseudocode, flow of scripts, and/or link to location of code used in analyses.
```{r analysis, echo = T, eval = F}
# Insert analysis code here
```



References
<!-- References must be submitted in BibTeX format. When compiling this document, they will appear in this section. There is no need to write them out here. Don't forget to include the name of your .bib file in the YAML header (fill in XYZ.bib in YAML header)!!!!!--> 






