# Annual SST Cycles

```{r setup,  echo = F, message=F}

#Load packages
library(knitr)
library(rmarkdown)

```


## Contributor name(s)
Sean Hardison, Vincent Saba

## Data name
Annual SST Cycles

### Indicator category 

<!-- 1. Database pull -->
2. Database pull with analysis
<!-- 3. Synthesis of published information -->
<!-- 4. Extensive analysis; not yet published -->
<!-- 5. Published methods -->

## Methods

### Data source(s) 
Data for annual sea surface tempature cycles were derived from the NOAA optimum interpolation sea surface temperature high resolution dataset ([NOAA OISST V2 dataset](https://www.esrl.noaa.gov/psd/data/gridded/data.noaa.oisst.v2.highres.html)). The data extend from 1981 to present, and provide a 0.25&deg; x 0.25&deg; global grid of SST measurements [@Reynolds2007]. Gridded SST data were masked according to the extent of Ecological Production Units in the Northeast Large Marine Ecosystem (NE-LME) (See ["EPU_Extended" shapefiles](https://github.com/NOAA-EDAB/tech-memo/tree/master/gis)).


### Data extraction 
Daily mean sea surface temperature data for 2017 and for each year during the period of 1981-2012 were downloaded from the NOAA [OI SST V5 site](https://www.esrl.noaa.gov/psd/cgi-bin/db_search/DBListFiles.pl?did=132&tid=69156&vid=2423) to derive the long-term climatological mean for the period. 

R Libraries Used:
```{r libraries, echo = T, eval = F}
#libraries
library(ncdf4);library(dplyr)
library(readr);library(tidyr)
library(sp);library(rgdal)
library(raster);library(stringr)

```

R code used in extraction and processing
```{r extraction, echo = T, eval = F}

#get spatial polygons for Ecological Production Units (EPUs) that are used to clip SST data.
EPU <- readOGR('Extended_EPU')

map.crs <- CRS("+proj=longlat +lat_1=35 +lat_2=45 +lat_0=40 +lon_0=-77 +x_0=0
               +y_0=0 +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0")

#find long term daily mean SSTs and 2017 SST anomaly
MAB_sst_daily_mean <- NULL
GB_sst_daily_mean <- NULL
GOM_sst_daily_mean <- NULL

#I split the data into three directories to loop through in separate R sessions concurrently. 
for (dir. in 1:3){
  
  #Loop through directories
  setwd(paste0('c:/users/sean.hardison/documents/sst_data/',dir.))
  print(getwd())
  
  for (f in 1:length(list.files())){
    
    if (!str_detect(list.files()[f],".nc")){
      print(paste(list.files()[f],"is not a raster")) #Based on file type
      next
    }
    
    for (j in c("MAB","GB","GOM")){
      
      sub_region <- EPU[EPU@data$EPU == j,]
      y <- as.numeric(str_extract(list.files()[f],"[0-9]+")) #get year
      
      for (i in 1:365){
        print(paste(j,y,i))
        daily_mean <- raster(paste0(list.files()[f]), band = i) #get band
        
        #set crs
        daily_mean@crs <- sub_region@proj4string 
        
        
        #rotate to lon scale from 0-360 to -180-180
        daily_mean <- rotate(daily_mean)
        
        #mask raster with spatialpolygon
        daily_mean_clipped <- mask(daily_mean, sub_region)
        
        
        #add mean value to data.frame
        assign(paste0(j,"_sst_daily_mean"),rbind(get(paste0(j,"_sst_daily_mean")),
                                                 c(mean(daily_mean_clipped@data@values, na.rm = T),y,i)))
        
      }
    }
    
    
  }
}

#Put results into data.frames
mab <- data.frame(EPU = "MAB",
                  year = MAB_sst_daily_mean[,2],
                  day = MAB_sst_daily_mean[,3],
                  Value = MAB_sst_daily_mean[,1])

gb <- data.frame(EPU = "GB",
                 year = GB_sst_daily_mean[,2],
                 day = GB_sst_daily_mean[,3],
                 Value = GB_sst_daily_mean[,1])

gom <- data.frame(EPU = "GOM",
                  year = GOM_sst_daily_mean[,2],
                  day = GOM_sst_daily_mean[,3],
                  Value = GOM_sst_daily_mean[,1])


data3 <- rbind(mab, gb, gom)

#Save as 1 of 3 files (one for each directory containing daily mean data)
save(data3, file = "dir3_sst.Rdata")

#--------------------------2017 SSTs----------------------------#
MAB_2017 <- NULL
GB_2017 <- NULL
GOM_2017 <- NULL

for (j in c("MAB","GB","GOM")){
  
  sub_region <- EPU[EPU@data$EPU == j,]
  
  for (i in 1:365){
    print(paste(j,i))
    daily_mean <- raster("sst.day.mean.2017.nc", band = i) #get band
    
    
    #set crs
    daily_mean@crs <- sub_region@proj4string 
    
    
    #rotate to lon scale from 0-360 to -180-180
    daily_mean <- rotate(daily_mean)
    
    #mask raster with spatialpolygon
    daily_mean_clipped <- mask(daily_mean, sub_region)
    
    
    #add mean value to data.frame
    assign(paste0(j,"_2017"),rbind(get(paste0(j,"_2017")),
                                             c(mean(daily_mean_clipped@data@values, na.rm = T),"2017",i)))
    
  }
}
#Put results into data.frames
mab_2017 <- data.frame(EPU = "MAB",
                  year = MAB_2017[,2],
                  day = MAB_2017[,3],
                  Value = MAB_2017[,1])

gb_2017 <- data.frame(EPU = "GB",
                 year = GB_2017[,2],
                 day = GB_2017[,3],
                 Value = GB_2017[,1])

gom_2017 <- data.frame(EPU = "GOM",
                  year = GOM_2017[,2],
                  day = GOM_2017[,3],
                  Value = GOM_2017[,1])


#Final 2017 daily mean data
sst_2017 <- rbind(mab_2017, gb_2017, gom_2017)
#save(sst_2017, file = "sst_2017.Rdata")

```

### Data analysis 
We found the long-term mean and standard deviation of SST over the period of 1982-2012 for each EPU, as well as the daily mean for 2017.  

```{r analyses, echo = T, eval = F}
#----------------------Load results--------------------------#
load("dir1_sst.Rdata")
load("dir2_sst.Rdata")
load("dir3_sst.Rdata")
load("sst_2017.Rdata")

#Get long term mean and standard deviation
d <- rbind(data1, data2, data3)

ltm <- d %>% group_by(EPU, day) %>% dplyr::summarise(mean  = mean(Value),
                                                     sd = sd(Value)) 

```


```{r plotting, echo = F, eval = T, fig.cap = "Long-term mean SSTs for the Mid-Atlantic Bight (A), Georges Bank (B), and Gulf of Maine (C) are shown by the black line, with +/- 1 SD shaded in grey. Red or dark blue shaded regions show where mean daily SSTs exceeded +/- 1 SD in 2017.", fig.width=8, fig.height=3.25, fig.align='center'}
data.dir <- "./data"
load(file.path(data.dir,"SOE_data_2018.Rdata"))

##---------------------------------MAB-----------------------------------------#
par( mfrow = c(1,3))
doy <- as.numeric(SOE.data.2018[SOE.data.2018$Var == "sst mean 2017 MAB",]$Time)
val_2017 <- SOE.data.2018[SOE.data.2018$Var == "sst mean 2017 MAB",]$Value
val_LT <- SOE.data.2018[SOE.data.2018$Var == "sst mean long term MAB",]$Value
val_LT_sd <- SOE.data.2018[SOE.data.2018$Var == "sst sd long term MAB",]$Value


# val_2017 <- approx(doy,val_2017, xout = seq(doy[1],doy[length(doy)],length.out = 365*1))$y
# val_LT <- approx(doy,val_LT, xout = seq(doy[1],doy[length(doy)],length.out = 365*1))$y
# val_LT_sd <- approx(doy,val_LT_sd, xout = seq(doy[1],doy[length(doy)],length.out = 365*1))$y
doy <- seq(doy[1],doy[length(doy)],length.out = 365*1)


above_mean <- NULL
for (i in 1:length(val_2017)){
  if (val_2017[i] >= val_LT[i]){
    above_mean[i] <- val_2017[i]
  } else if (val_2017[i] < val_LT [i]){
    above_mean[i] <- NA
  }
}

below_mean <- NULL
for (i in 1:length(val_2017)){
  if (val_2017[i] <= val_LT[i]){
    below_mean[i] <- val_2017[i]
  } else if (val_2017[i] > val_LT [i]){
    below_mean[i] <- NA
  }
}

above_sd <- NULL
for (i in 1:length(val_2017)){
  if (val_2017[i] >= val_LT_sd[i] + val_LT[i]){
    above_sd[i] <- val_2017[i]
  } else if (val_2017[i] < val_LT_sd [i] + val_LT[i]){
    above_sd[i] <- NA
  }
}

below_sd <- NULL
for (i in 1:length(val_2017)){
  if (val_2017[i] <= val_LT[i] - val_LT_sd[i]){
    below_sd[i] <- val_2017[i]
  } else if (val_2017[i] > val_LT[i] - val_LT_sd [i]){
    below_sd[i] <- NA
  }
}

#Lines for polygons
above_sd[is.na(above_sd)] <- val_LT_sd[which(is.na(above_sd))] + val_LT[which(is.na(above_sd))]
below_sd[is.na(above_sd)] <- val_LT[which(is.na(below_sd))] - val_LT_sd[which(is.na(below_sd))] 
above_mean[is.na(above_mean)] <- val_LT[which(is.na(above_mean))]
below_mean[is.na(below_mean)] <- val_LT[which(is.na(below_mean))]

upper <- val_LT_sd + val_LT
lower <- val_LT - val_LT_sd

#Null figure
plot(NULL, xlim = c(doy[1],doy[(length(doy))]), ylim = c(4,25), las = 1, 
     ylab = "", yaxt = "n", xaxt = "n", xlab = "")
axis(2, cex.axis = 1.25, las = 1)
axis(1,  labels = c("Jan","Mar","May","July","Sep","Nov","Jan"), 
     at = c(1,61,122,183,245,306,365), cex.axis= 1.25)
mtext(2, line = 2.3, text = expression(paste("Mean SST (",degree,"C)")), cex = 1.1)
mtext(1, line = 2.5, text = "Time", cex = 1.1)
text(15,25*.95,"A",cex = 1.5)
# +/- 1 sd
polygon(c(doy, rev(doy)),
        c(upper, rev(lower)),
        col = "grey85", border = NA)

#Fills plot
polygon(c(doy, rev(doy)),
        c(below_mean + (val_LT-below_mean), rev(below_mean)),
        col = "lightblue", border = NA)
polygon(c(doy, rev(doy)),
        c(above_mean - (above_mean-val_LT), rev(above_mean)),
        col = "orange", border = NA)
polygon(c(doy, rev(doy)),
        c(above_sd - (above_sd-(val_LT + val_LT_sd)), rev(above_sd)),
        col = "red", border = NA)
polygon(c(doy, rev(doy)),
        c(below_sd + (below_sd-(val_LT - val_LT_sd)), rev(below_sd)),
        col = "blue", border = NA)
points(doy,val_LT, type = "l", lwd = 1, "grey90")


##-------------------------------------GB-------------------------------------#

doy <- as.numeric(SOE.data.2018[SOE.data.2018$Var == "sst mean 2017 GB",]$Time)
val_2017 <- SOE.data.2018[SOE.data.2018$Var == "sst mean 2017 GB",]$Value
val_LT <- SOE.data.2018[SOE.data.2018$Var == "sst mean long term GB",]$Value
val_LT_sd <- SOE.data.2018[SOE.data.2018$Var == "sst sd long term GB",]$Value


# val_2017 <- approx(doy,val_2017, xout = seq(doy[1],doy[length(doy)],length.out = 365*1))$y
# val_LT <- approx(doy,val_LT, xout = seq(doy[1],doy[length(doy)],length.out = 365*1))$y
# val_LT_sd <- approx(doy,val_LT_sd, xout = seq(doy[1],doy[length(doy)],length.out = 365*1))$y
doy <- seq(doy[1],doy[length(doy)],length.out = 365*1)


above_mean <- NULL
for (i in 1:length(val_2017)){
  if (val_2017[i] >= val_LT[i]){
    above_mean[i] <- val_2017[i]
  } else if (val_2017[i] < val_LT [i]){
    above_mean[i] <- NA
  }
}

below_mean <- NULL
for (i in 1:length(val_2017)){
  if (val_2017[i] <= val_LT[i]){
    below_mean[i] <- val_2017[i]
  } else if (val_2017[i] > val_LT [i]){
    below_mean[i] <- NA
  }
}

above_sd <- NULL
for (i in 1:length(val_2017)){
  if (val_2017[i] >= val_LT_sd[i] + val_LT[i]){
    above_sd[i] <- val_2017[i]
  } else if (val_2017[i] < val_LT_sd [i] + val_LT[i]){
    above_sd[i] <- NA
  }
}

below_sd <- NULL
for (i in 1:length(val_2017)){
  if (val_2017[i] <= val_LT[i] - val_LT_sd[i]){
    below_sd[i] <- val_2017[i]
  } else if (val_2017[i] > val_LT[i] - val_LT_sd [i]){
    below_sd[i] <- NA
  }
}

#Lines for polygons
above_sd[is.na(above_sd)] <- val_LT_sd[which(is.na(above_sd))] + val_LT[which(is.na(above_sd))]
below_sd[is.na(above_sd)] <- val_LT[which(is.na(below_sd))] - val_LT_sd[which(is.na(below_sd))] 
above_mean[is.na(above_mean)] <- val_LT[which(is.na(above_mean))]
below_mean[is.na(below_mean)] <- val_LT[which(is.na(below_mean))]

upper <- val_LT_sd + val_LT
lower <- val_LT - val_LT_sd

#Null figure
plot(NULL, xlim = c(doy[1],doy[(length(doy))]), ylim = c(4,21), las = 1, 
     ylab = "", yaxt = "n", xaxt = "n", xlab = "")
axis(2, cex.axis = 1.25, las = 1)
axis(1,  labels = c("Jan","Mar","May","July","Sep","Nov","Jan"), 
     at = c(1,61,122,183,245,306,365), cex.axis= 1.25)
#mtext(2, line = 2.5, text = expression(paste("Mean SST (",degree,"C)")), cex = 1.1)
mtext(1, line = 2.5, text = "Time", cex = 1.1)
text(15,21*.95,"B",cex = 1.5)
# +/- 1 sd
polygon(c(doy, rev(doy)),
        c(upper, rev(lower)),
        col = "grey85", border = NA)

#Fills plot
polygon(c(doy, rev(doy)),
        c(below_mean + (val_LT-below_mean), rev(below_mean)),
        col = "lightblue", border = NA)
polygon(c(doy, rev(doy)),
        c(above_mean - (above_mean-val_LT), rev(above_mean)),
        col = "orange", border = NA)
polygon(c(doy, rev(doy)),
        c(above_sd - (above_sd-(val_LT + val_LT_sd)), rev(above_sd)),
        col = "red", border = NA)
polygon(c(doy, rev(doy)),
        c(below_sd + (below_sd-(val_LT - val_LT_sd)), rev(below_sd)),
        col = "blue", border = NA)
points(doy,val_LT, type = "l", lwd = 1, "grey90")

#----------------------------------------------------------------------------#

## SST GOM

doy <- as.numeric(SOE.data.2018[SOE.data.2018$Var == "sst mean 2017 GOM",]$Time)
val_2017 <- SOE.data.2018[SOE.data.2018$Var == "sst mean 2017 GOM",]$Value
val_LT <- SOE.data.2018[SOE.data.2018$Var == "sst mean long term GOM",]$Value
val_LT_sd <- SOE.data.2018[SOE.data.2018$Var == "sst sd long term GOM",]$Value


# val_2017 <- approx(doy,val_2017, xout = seq(doy[1],doy[length(doy)],length.out = 365*1))$y
# val_LT <- approx(doy,val_LT, xout = seq(doy[1],doy[length(doy)],length.out = 365*1))$y
# val_LT_sd <- approx(doy,val_LT_sd, xout = seq(doy[1],doy[length(doy)],length.out = 365*1))$y
doy <- seq(doy[1],doy[length(doy)],length.out = 365*1)


above_mean <- NULL
for (i in 1:length(val_2017)){
  if (val_2017[i] >= val_LT[i]){
    above_mean[i] <- val_2017[i]
  } else if (val_2017[i] < val_LT [i]){
    above_mean[i] <- NA
  }
}

below_mean <- NULL
for (i in 1:length(val_2017)){
  if (val_2017[i] <= val_LT[i]){
    below_mean[i] <- val_2017[i]
  } else if (val_2017[i] > val_LT [i]){
    below_mean[i] <- NA
  }
}

above_sd <- NULL
for (i in 1:length(val_2017)){
  if (val_2017[i] >= val_LT_sd[i] + val_LT[i]){
    above_sd[i] <- val_2017[i]
  } else if (val_2017[i] < val_LT_sd [i] + val_LT[i]){
    above_sd[i] <- NA
  }
}

below_sd <- NULL
for (i in 1:length(val_2017)){
  if (val_2017[i] <= val_LT[i] - val_LT_sd[i]){
    below_sd[i] <- val_2017[i]
  } else if (val_2017[i] > val_LT[i] - val_LT_sd [i]){
    below_sd[i] <- NA
  }
}

#Lines for polygons
above_sd[is.na(above_sd)] <- val_LT_sd[which(is.na(above_sd))] + val_LT[which(is.na(above_sd))]
below_sd[is.na(above_sd)] <- val_LT[which(is.na(below_sd))] - val_LT_sd[which(is.na(below_sd))] 
above_mean[is.na(above_mean)] <- val_LT[which(is.na(above_mean))]
below_mean[is.na(below_mean)] <- val_LT[which(is.na(below_mean))]

upper <- val_LT_sd + val_LT
lower <- val_LT - val_LT_sd

#Null figure
plot(NULL, xlim = c(doy[1],doy[(length(doy))]), ylim = c(4,21), las = 1, 
     ylab = "", yaxt = "n", xaxt = "n", xlab = "")
axis(2, cex.axis = 1.25, las = 1)
axis(1,  labels = c("Jan","Mar","May","July","Sep","Nov","Jan"), 
     at = c(1,61,122,183,245,306,365), cex.axis= 1.25)
#(2, line = 2.5, text = expression(paste("Mean SST (",degree,"C)")), cex = 1.1)
mtext(1, line = 2.5, text = "Time", cex = 1.1)
text(15,21*.95,"C",cex = 1.5)
# +/- 1 sd
polygon(c(doy, rev(doy)),
        c(upper, rev(lower)),
        col = "grey85", border = NA)

#Fills plot
polygon(c(doy, rev(doy)),
        c(below_mean + (val_LT-below_mean), rev(below_mean)),
        col = "lightblue", border = NA)
polygon(c(doy, rev(doy)),
        c(above_mean - (above_mean-val_LT), rev(above_mean)),
        col = "orange", border = NA)
polygon(c(doy, rev(doy)),
        c(above_sd - (above_sd-(val_LT + val_LT_sd)), rev(above_sd)),
        col = "red", border = NA)
polygon(c(doy, rev(doy)),
        c(below_sd + (below_sd-(val_LT - val_LT_sd)), rev(below_sd)),
        col = "blue", border = NA)
points(doy,val_LT, type = "l", lwd = 1, "grey90")
box()
```



### Further metadata

#### Public availability statement
Source data are available [here](https://www.esrl.noaa.gov/psd/cgi-bin/db_search/DBListFiles.pl?did=132&tid=69156&vid=2423)


#### Point of contact
Sean Hardison

#### Data steward
Sean Hardison

<!-- References must be submitted in BibTeX format. When compiling this document, they will appear in this section. There is no need to write them out here. Don't forget to include the name of your .bib file in the YAML header (fill in XYZ.bib in YAML header)!!!!!--> 






