# Aquaculture

```{r setup,  echo = F, message=F}

#Load packages
library(knitr)
library(rmarkdown)

```
**Description**: Aquaculture indicators

**Indicator category**: Synthesis of published information

**Contributor(s)**: Sean Hardison, Lisa Calvo, Karl Roscher
  
**Data steward**: Sean Hardison <sean.hardison@noaa.gov>
  
**Point of contact**: Sean Hardison <sean.hardison@noaa.gov> 
  
**Public availability statement**: Source data are publicly available in referenced reports, and are also available for download [here](https://comet.nefsc.noaa.gov/erddap/tabledap/aquaculture_2018.html). 


## Methods
Aquaculture data included in the SOE report were time series of number of oysters sold in Virginia, Maryland, and New Jersey. 


### Data sources
Virginia oyster harvest data are collected from mail and internet-based surveys of active oyster aquaculture operations on both sides of the Chesapeake Bay, which are then synthesized in an annual report [@Hudson2017a]. In Maryland, shellfish aquaculturists are required to report their monthly harvests to the Maryland Department of Natural Resources (MD-DNR). The MD-DNR then aggregates the harvest data for release in the Maryland Aquaculture Coordinating Council Annual Report [@ACC2017], from which data were collected. Similar to Virginia, New Jersey releases annual reports synthesizing electronic survey results from lease-holding shellfish growers. Data from New Jersey reflects cage reared oysters grown from hatchery seed [@Calvo2017]. 


### Data extraction 
Data were collected directly from state aquaculture reports. Oysters harvest data in MD was reported in bushels which were then converted to individual oysters by an estimate of 300 oysters bushel$^{-1}$.

### Data analysis
No data analyses occurred for this indicator.

### Plotting

```{r, mariculture, fig.cap="Oyster aquaculture production in terms of number of oysters sold from Virginia, Maryland, and New Jersey.",echo = T, warning = F, message=F, fig.align='center', fig.height=4}

library(dplyr)

# Relative working directories
data.dir  <- here::here('data')
r.dir <- here::here('R')

# Load data
load(file.path(data.dir,"SOE_data_2018_erddap.Rdata"))

#colors
topo <- c('black', "indianred", "darkorange")

#filter
oyst_vars <- SOE.data.2018 %>%
  dplyr::filter(Var %in%
          c("aquaculture NJ gear raised oysters sold",
            "aquaculture MD oyster harvest N",
            "aquaculture VA oyster harvest N")) %>% 
  tidyr::spread(.,Var,Value)

#plotting
opar <- par(mar = c(4, 6, 2, 6))
y.min <- 1
y.max <- 40e6
time <- oyst_vars$Time
plot(NULL, xlim = c(2005, 2017),
     ylim = c(y.min,y.max), xlab = '', ylab = '', yaxt = "n",ty = 'n',
     cex.lab=1, las = 1, cex.axis = 1)
points(time, oyst_vars$`aquaculture NJ gear raised oysters sold`, type = "o", lwd = 2.5, col = topo[1], pch = 20)
points(time, oyst_vars$`aquaculture MD oyster harvest N`, type = "o", lwd = 2.5, col = topo[2], pch = 20)
points(time, oyst_vars$`aquaculture VA oyster harvest N`, type = "o", lwd = 2.5, col = topo[3], pch = 20)

mtext(1, text = 'Year', line = 2)
mtext(2, text = expression("Oysters Harvested, 10"^6*'n'), line = 2.5)
axis(2, at = c(1*10^6,5*10^6,10*10^6,15*10^6,
                 20*10^6,25*10^6,30*10^6,35*10^6,40*10^6), 
       labels = c("1","5","10","15","20",
                  "25","30","35","40"),
       las = 1, cex.axis = 1)
legend(2006, 40e6, legend = c("New Jersey",
                            "Maryland",
                            "Virginia"), col = topo, bty = "n",lwd = 2)

```




