# Fishery Reliance and Social Vulnerability

```{r setup,  echo = F, message=F}

#Load packages
library(knitr)
library(rmarkdown)

```
**Description**: Fishing community commercial and recreational fishing reliance and social vulnerability

**Indicator category**: Database pull with analysis

**Contributor(s)**: Lisa L. Colburn
  
**Data steward**: Lisa L. Colburn
  
**Point of contact**: Lisa L. Colburn 
  
**Public availability statement**: The source data used to construct the commercial fishing engagement and reliance indices include confidential information and are not available publicly. However, the commercial fishing engagement and reliance indices are not confidential so are available to the public. All calculated indices can be found [here](https://www.st.nmfs.noaa.gov/humandimensions/social-indicators/map).
  

## Methods


### Data sources
NOAA Fisheries' Community Social Vulnerability Indicators (CSVIs) were developed using secondary data including social, demographic and fisheries variables. The social and demographic data were downloaded from the 2014 American Community Survey (ACS) 5-yr estimates Dataset at the [U.S. Census American FactFinder site](https://www.census.gov/programs-surveys/acs/) for coastal communities at the Census Designated Place (CDP) level, and in some cases the County Subdivision (MCD) level. Commercial fisheries data were pulled from the SOLE server located at Northeast Fisheries Science Center in Woods Hole, MA. The recreational fishing information is publicly accessible through the [Marine Recreational Information Program (MRIP)](https://www.st.nmfs.noaa.gov/recreational-fisheries/MRIP/), and for this analysis was custom requested from NOAA Fisheries headquarters.


### Data extraction 
Commercial fisheries data was pulled from the NEFSC SOLE server in Woods Hole, MA.

SQL and SAS code for data extraction and processing steps:
```{sql, eval = F, echo = T}
create table cfders2011 as
select *
from connection to oracle
(select port, state, year, dealnum, permit, nespp3, spplndlb, sppvalue from cfders2011 where permit > 0 order by permit);
create table cfvess11 as
select *
from connection to oracle
(select permit, homeport, homest from CFDBS.cfvess11 where permit > 0 order by permit);
create table port_name as
select *
from connection to oracle
(select port, portnm from port order by port);
create table st_name as
select *
from connection to oracle
(select state, stateabb from statenm order by state);


Truncated SAS code:

/*CREATE VARIABLES FOR TOTAL LANDINGS WEIGTH AND VALUE (SUM) BY PORT OF LANDING AND BY HOMEPORT*/
data landings_ports1; set landings_ports;
run;
proc sort;
by port state;
run;
proc means noprint data = landings_ports1; by port state; 
var spplndlb sppvalue;
id port state;
output out = landport_totspp sum = L_Totlb L_Totval;
run;
proc sort;
by port;
run;
data landings_ports2; set landings_ports;
run;
proc sort;
by homeport homest;
run;
proc means noprint data = landings_ports2; by homeport homest; 
var spplndlb sppvalue;
id homeport homest;
output out = homeport_totspp sum = H_Totlb H_Totval;
run;
proc sort;
by homeport;
run;

/*CREATE SPECIES VARIABLES*/
data landings_ports_NE_spp; set landings_ports;
monklb = 0; monkval = 0; /*monkfish*/
bluelb = 0; blueval = 0; /*bluefish*/
.omitted.
otherlb = 0; otherval = 0; /*other - everything else*/
run;
data landings_ports_NE_spp2; set landings_ports_NE_spp;
if nespp3 = 012 then do; monklb = spplndlb; monkval = sppvalue; end;
...ommitted.
if nespp3 = 406 then do; spotlb = spplndlb; spotval = sppvalue; end;
if nespp3 not in (012, 023, 033, 051, 081, 105, 112, 115, 116, 120, 121, 122, 123, 124, 125, 132, 147, 152, 153, 155, 159, 168, 194, 197, 212, 
221, 240, 250, 269, 305, 329, 330, 335, 344, 345, 351, 352, 365, 366, 367, 368, 369, 370, 372, 373, 384, 415, 418, 432, 438, 443, 444, 445, 
446, 447, 464, 466, 467, 468, 469, 470, 471, 472, 507, 508, 509, 512, 517, 700, 710, 711, 724, 727, 748, 754, 769, 774, 775, 781, 786, 789, 
798, 799, 800, 801, 802, 805, 806, 899, 001, 090, 069, 107, 150, 173, 196, 334, 347, 349, 364, 371, 420, 422, 481, 484, 714, 776, 777, 823, 763, 736) 
then do; otherlb = spplndlb; otherval = sppvalue; end;
run; 

/*SUM SPECIES LANDINGS BY PORT OF LANDING*/
proc sort; by port; proc means noprint data = landings_ports_NE_spp2; by port state; 
. omitted ...
id port state;
output out = spp_porlnd_NE sum = ;
run;
proc sort;
by port;
run;

/*SUM SPECIES LANDINGS BY HOMEPORT*/
data spp_home; set landings_ports_NE_spp2;
run;
proc sort; by homeport homest; proc means noprint data = spp_home; by homeport homest; 
. species are counted..
id homeport homest;
output out = spp_homep_NE sum = ;
run
proc sort;
by homeport; run;

/*MERGE TOTAL PERMITS AND TOTAL DEALERS BY PORT OF LANDING*/
data land_port_totperm2; set land_port_totperm;
run;
proc sort;
by port; run;
data lnd_port_permit; merge spp_porlnd_NE (IN=X) land_port_totperm2 (IN=Y);
by port; if X=1; run;data land_port_totdeal2; set land_port_totdeal;
run;
proc sort;
by port;
run;
data lnd_port_permit_deal; merge lnd_port_permit (IN=x) land_port_totdeal2 (IN=Y);
by port; if X=1; run;

/*MERGE WITH PORT NAME AND STATE ABBREVIATION*/
data lnd_port_permit_deal_nm; merge lnd_port_permit_deal (IN=X) port_name (IN=Y);
by port; if X=1; run;
data lnd_port_permit_deal_nm_st; merge lnd_port_permit_deal_nm (IN=x) st_name (IN=Y); proc sort;
by port; if X=1; run;

/*MERGE TOTAL PERMITS AND TOTAL DEALERS BY HOMEPORT*/
data home_port_totperm2; set home_port_totperm;
run;
proc sort;
by homeport;
run;
data home_port_permit; merge spp_homep_NE (IN=X) home_port_totperm2 (IN=Y);
by homeport; if X=1; run; data home_port_totdeal2; set home_port_totdeal;
run;
proc sort;
by homeport;
run;
data home_port_permit_deal; merge home_port_permit (IN=x) home_port_totdeal2 (IN=Y);
by homeport; if X=1; run; proc sort;
by homeport;
run;

/*MERGE TOTAL LANDINGS BY PORT OF LANDING*/
data lnd_port_per_deal_nm_st_tspp; merge lnd_port_permit_deal_nm_st (IN=X) landport_totspp (IN=Y);
by port; if X=1; run;

/*MERGE TOTAL LANDINGS BY HOMEPORT*/
data home_port_per_deal_tspp; merge home_port_permit_deal (IN=X) homeport_totspp (IN=Y);
by homeport; if X=1; run;
data netana.port_landing11; set lnd_port_per_deal_nm_st_tspp;
if state in (22, 32, 24, 42, 7, 35, 33, 8, 23, 49, 36);
run;
proc sort;
by port state;
run;
data netana.homeport11; set home_port_per_deal_tspp;
if homest in ('ME', 'NH', 'MA', 'RI', 'CT', 'NY', 'NJ', 'DE', 'MD', 'VA', 'NC');
run;
proc sort;
by homeport homest;
run;


```


### Data analysis
The indicators were developed using the methodology described in @Jacob2010, @Jacob2013, @colburn_social_2012 and @jepson_development_2013. Indicators were constructed through principal component analysis with a single factor solution, and the following criteria had to have been met: a minimum variance explained of 45%; Kasier-Meyer Olkin measure of sampling adequacy above.500; factor loadings above.350; Bartlett's test of sphericity significance above .05; and an Armor's Theta reliability coefficient above .500. Factor scores for each community were ranked based on standard deviations into the following categories: High(>=1.00SD), MedHigh .500-.999 SD), Moderate (.000-.499 SD) and Low (<.000 SD).

### Plotting

```{r map1, echo = T, eval=T, fig.cap="Commercial (A) and recreational (B) reliance and social vulnerability in the Mid-Atlantic Bight",  fig.width = 8, fig.height = 5.4, fig.align='center', fig.show='hold', message=F, warning=F}
  
#Libraries
library(rgdal)
library(maps);library(raster);
library(mapdata);library(grid);
library(stringr);library(png)
library(ncdf4);library(dplyr)

#Get data and set projection
data.dir <- here::here("data")
gis.dir <- here::here("gis")


load(file.path(data.dir, "SOE_data_erddap.Rdata"))

#coastline
coast <- readOGR(gis.dir, 'NES_LME_coast', verbose = F)


#projection
map.crs <- CRS("+proj=longlat +lat_1=35 +lat_2=45 +lat_0=40 +lon_0=-77 +x_0=0
               +y_0=0 +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0")
coast <- spTransform(coast,map.crs)



#Make figures
par(mfrow=c(1,2), mar = c(2,2.85,1,1))

cat1 <- SOE.data[SOE.data$Var == "social vulnerability MAB",]$Value
cat2 <- SOE.data[SOE.data$Var == "commercial reliance MAB",]$Value

#group data into categories
ncat1 <- NULL
for (i in 1:length(cat1)){
  if (cat1[i] > 2){
    ncat1[i] <- "C"
  } else if ((cat1[i] <= 2) & (cat1[i] > 1)){
    ncat1[i] <- "B"
  } else {
    ncat1[i] <- "A"
  }
}

ncat2 <- NULL
for (i in 1:length(cat1)){
  if (cat2[i] > 2){
    ncat2[i] <- 3
  } else if ((cat2[i] <= 2) & (cat2[i] > 1)){
    ncat2[i] <- 2
  } else {
    ncat2[i] <- 1
  }
}


cat3 <- paste0(ncat1, ncat2)
cat3 <- factor(cat3, levels = c("A1","A2","A3",
                                "B1","B2","B3",
                                "C1","C2","C3"),
               ordered = TRUE)
lon <- SOE.data[SOE.data$Var == "choropleth longitude MAB",]$Value
lat <- SOE.data[SOE.data$Var == "choropleth latitude MAB",]$Value

#new dataframe to turn into sp object
mab.dat <- data.frame(lon = lon,
                      lat = lat,
                      cat3 = cat3)

#break up into groups for plotting - allows for effective layering
g1 <- mab.dat %>% filter(cat3 == "A1"|cat3 == "A2"|cat3 == "A3") %>% arrange(cat3)
g2 <- mab.dat %>% filter(cat3 == "B1"|cat3 == "B2") %>% arrange(cat3)
g3 <- mab.dat %>% filter(cat3 == "C1") %>% arrange(cat3)


g2_big <- mab.dat %>% filter(cat3 == "B3")
g3_big <- mab.dat %>% filter(cat3 == "C2"|cat3 == "C3") %>% arrange(cat3)


#turn grouped data into sp objects for overlaying on plot of Mid-Atlantic

coordinates(g1) <- ~lon+lat
g1@proj4string <- map.crs
g1 <- spTransform(g1, map.crs)
colo1 <- c(rep("#FFFFFF",table(g1$cat3)[1]),
           rep("#B8B8FF",table(g1$cat3)[2]),
           rep("#7F7FFF",table(g1$cat3)[3]))


coordinates(g2) <- ~lon+lat
g2@proj4string <- map.crs
g2 <- spTransform(g2, map.crs)
colo2 <- c(rep("#FFB8B8",table(g2$cat3)[4]),rep("#B871B8",table(g2$cat3)[5]))


coordinates(g3) <- ~lon+lat
g3@proj4string <- map.crs
g3 <- spTransform(g3, map.crs)
colo3 <- c("#FF7F7F")


coordinates(g2_big) <- ~lon+lat
g2_big@proj4string <- map.crs
g2_big <- spTransform(g2_big, map.crs)

coordinates(g3_big) <- ~lon+lat
g3_big@proj4string <- map.crs
g3_big <- spTransform(g3_big, map.crs)


#plot map and dots of different size based on category

plot(coast, xlim = c(-77.5,-71), ylim = c(36.5,41),col = "grey")
plot(g1, pch = 16, col = colo1,cex = 1, add = T)
plot(g2, pch = 16, col = colo2,cex = 1, add = T)
plot(g3, pch = 16, col = colo3,cex = 1,add = T)


plot(g2_big, pch = 16, col = "blue",cex = 2, add = T)
plot(g3_big, pch = 16, col = c(rep("red",table(g3_big$cat3)[8]),rep("black",table(g3_big$cat3)[9])),cex = 2,add = T)

#colors of bivariate color scheme
base_col <- c("#7F7FFF","blue","black",
              "#B8B8FF","#B871B8","red",
              "#FFFFFF","#FFB8B8","#FF7F7F")

#raster image to display on plot as bivariate legend
r <- raster(xmn = -73.5, xmx = -71.4, ymn = 36, ymx = 37.8, nrows = 3, ncols = 3)
r[] <- 1:9
plot(r, col = base_col, legend = F, bty = "n", axes = F, yaxt = "n", frame.plot = F,
     xaxt='n', add = T)

#text
text(-73.9, 36.9, "Commercial Reliance",srt=90, cex = .75)
text(-72.4,35.6, "Social Vulnerability", cex = .75)
text(-77.1,42, "A", cex = 1.5)

axis(1, at = c(-76,-74,-72), labels = paste( c(-76,-74,-72) * -1, 'W'),pos = 34.7,col = NA, col.ticks = 1)
axis(2, at = axTicks(2), labels = paste(axTicks(2), 'N'), las = T, pos = -77.8,col = NA, col.ticks = 1)
box(lty = 1, lwd = 2)
#text

#####################################################################################################

cat1 <- SOE.data[SOE.data$Var == "social vulnerability MAB",]$Value
cat2 <- SOE.data[SOE.data$Var == "recreational reliance MAB",]$Value

#group data into categories
ncat1 <- NULL
for (i in 1:length(cat1)){
  if (cat1[i] > 2){
    ncat1[i] <- "C"
  } else if ((cat1[i] <= 2) & (cat1[i] > 1)){
    ncat1[i] <- "B"
  } else {
    ncat1[i] <- "A"
  }
}

ncat2 <- NULL
for (i in 1:length(cat1)){
  if (cat2[i] > 2){
    ncat2[i] <- 3
  } else if ((cat2[i] <= 2) & (cat2[i] > 1)){
    ncat2[i] <- 2
  } else {
    ncat2[i] <- 1
  }
}


cat3 <- paste0(ncat1, ncat2)
cat3 <- factor(cat3, levels = c("A1","A2","A3",
                                "B1","B2","B3",
                                "C1","C2","C3"),
               ordered = TRUE)
lon <- SOE.data[SOE.data$Var == "choropleth longitude MAB",]$Value
lat <- SOE.data[SOE.data$Var == "choropleth latitude MAB",]$Value

#new dataframe to turn into sp object
mab.dat <- data.frame(lon = lon,
                      lat = lat,
                      cat3 = cat3)

#break up into groups for plotting - allows for effective layering
g1 <- mab.dat %>% filter(cat3 == "A1"|cat3 == "A2"|cat3 == "A3") %>% arrange(cat3)
g2 <- mab.dat %>% filter(cat3 == "B1"|cat3 == "B2") %>% arrange(cat3)
g3 <- mab.dat %>% filter(cat3 == "C1") %>% arrange(cat3)


g2_big <- mab.dat %>% filter(cat3 == "B3")
g3_big <- mab.dat %>% filter(cat3 == "C2"|cat3 == "C3") %>% arrange(cat3)


#turn grouped data into sp objects for overlaying on plot of Mid-Atlantic

coordinates(g1) <- ~lon+lat
g1@proj4string <- map.crs
g1 <- spTransform(g1, map.crs)
colo1 <- c(rep("#FFFFFF",table(g1$cat3)[1]),
           rep("#B8B8FF",table(g1$cat3)[2]),
           rep("#7F7FFF",table(g1$cat3)[3]))


coordinates(g2) <- ~lon+lat
g2@proj4string <- map.crs
g2 <- spTransform(g2, map.crs)
colo2 <- c(rep("#FFB8B8",table(g2$cat3)[4]),rep("#B871B8",table(g2$cat3)[5]))


coordinates(g3) <- ~lon+lat
g3@proj4string <- map.crs
g3 <- spTransform(g3, map.crs)
colo3 <- c("#FF7F7F")


coordinates(g2_big) <- ~lon+lat
g2_big@proj4string <- map.crs
g2_big <- spTransform(g2_big, map.crs)

coordinates(g3_big) <- ~lon+lat
g3_big@proj4string <- map.crs
g3_big <- spTransform(g3_big, map.crs)


#plot map and dots of different size based on category

plot(coast, xlim = c(-77.5,-71), ylim = c(36.5,41),col = "grey")
plot(g1, pch = 16, col = colo1,cex = 1, add = T)
plot(g2, pch = 16, col = colo2,cex = 1, add = T)
plot(g3, pch = 16, col = colo3,cex = 1,add = T)


plot(g2_big, pch = 16, col = "blue",cex = 2, add = T)
plot(g3_big, pch = 16, col = c(rep("red",table(g3_big$cat3)[8]),rep("black",table(g3_big$cat3)[9])),cex = 2,add = T)

#colors of bivariate color scheme
base_col <- c("#7F7FFF","blue","black",
              "#B8B8FF","#B871B8","red",
              "#FFFFFF","#FFB8B8","#FF7F7F")

#raster image to display on plot as bivariate legend
r <- raster(xmn = -73.5, xmx = -71.4, ymn = 36, ymx = 37.8, nrows = 3, ncols = 3)
r[] <- 1:9
plot(r, col = base_col, legend = F, bty = "n", axes = F, yaxt = "n", frame.plot = F,
     xaxt='n', add = T)
axis(1, at = c(-76,-74,-72), labels = paste( c(-76,-74,-72) * -1, 'W'),pos = 34.7,col = NA, col.ticks = 1)
axis(2, at = axTicks(2), labels = paste(axTicks(2), 'N'), las = T, pos = -77.8,col = NA, col.ticks = 1)
box(lty = 1, lwd = 2)
#text
text(-73.9, 36.9, "Recreational Reliance",srt=90, cex = .75)
text(-72.4,35.6, "Social Vulnerability", cex = .75)
text(-77.1,42, "B", cex = 1.5)
```







