# Data and Code Access {#erddap}

```{r setup,  echo = F, message=F}

#Load packages
library(knitr)
library(rmarkdown)


```


### About

The Technical Documentation for the State of the Ecosystem reports is a [bookdown](https://bookdown.org) document; hosted on the NOAA Northeast Fisheries Science Center Ecosystems Dynamics and Assessment Branch [Github page](https://github.com/NOAA-EDAB), and developed in R. Derived data used to populate figures in this document are queried directly from the NEFSC [ERDDAP server](https://comet.nefsc.noaa.gov/erddap/info/index.html?page=1&itemsPerPage=1000) using the R package [rerddap](https://cran.r-project.org/web/packages/rerddap/vignettes/Using_rerddap.html).  

### Accessing source data and build code

Source data for the derived indicators in this document are linked to in the text unless there are privacy concerns involved. In that case, it may be possible to access source data by reaching out to the Point of Contact associated with that data set. Derived data sets make up the majority of the indicators present in the State of the Ecosystem reports. The majority of derived data sets included in State of the Ecosystem reports are available for download through the NEFSC [ERDDAP server](https://comet.nefsc.noaa.gov/erddap/info/index.html?page=1&itemsPerPage=1000), although some data is also present in the [Github repository](https://github.com/NOAA-EDAB/tech-doc) associated with this project. 

In this technical documentation, we hope to shine a light on the processing and analytical steps involved to get from source data to final product. This means that whenever possible, we have included the code involved in source data extraction, processing, and analyses. We have also attempted to thoroughly describe all methods in place of or in supplement to provided code. 

There are multiple R scripts sourced throughout this document in an attempt to keep code concise. These scripts include [BasePlot_source.R](https://github.com/NOAA-EDAB/tech-doc/blob/master/R/BasePlot_source.R), [GIS_source.R](https://github.com/NOAA-EDAB/tech-doc/blob/master/R/GIS_source.R), and [get_erddap.R](https://github.com/NOAA-EDAB/tech-doc/blob/master/R/get_erddap.R). These scripts along with the code presented in sections titled "Plotting" may be used to recreate the majority of figures present in State of the Ecosystem reports. The script [get_erddap.R](https://github.com/NOAA-EDAB/tech-doc/blob/master/R/get_erddap.R) is used to both query individual data sets and to build a parent data set containing all indicator data. The exact usage in the latter case is shown in the R code block below. For querying individual data sets using get_erddap.R, see the [single-species status indicator](#stockstatus).

### Building the document

Start a local build of the SOE bookdown document by first cloning the project's associated [git repository](https://github.com/NOAA-EDAB/tech-doc). Next, if you would like to build a past version of the document, use `git checkout [version_commit_hash]` to revert the project to a past commit of interest after cloning the repository, and set `build_latest <- FALSE` in the following code chunk. This will ensure the project builds from a cached data set, and not the most updated versions present on the NEFSC ERDDAP server. Once the `tech-doc.Rproj` file is opened in RStudio, run `bookdown::serve_book()` from the console to build the document. 

```{r query, echo = T, eval = T, message=F, warning=F}

build_latest <- FALSE

if (build_latest){
  #Libraries
  library(dplyr)
  library(rerddap)
  library(stringr)
  library(data.table)
  
  # Relative working directories
  data.dir  <- here::here('data')
  r.dir <- here::here('R')
  
  #Source function for querying ERDDAP server
  source(file.path(r.dir,"get_erddap.R"))
  
  #Set URL for COMET (server where NEFSC ERDDAP lives)
  comet <- 'https://comet.nefsc.noaa.gov/erddap/'
  
  #List datasets on the NEFSC ERDDAP
  tab_list <- ed_datasets(url = comet)
  
  #Get updated data set IDs
  erddap_datasets <- tab_list %>% 
    filter(str_detect(Dataset.ID, "soe_v")) %>% 
    get_erddap(id = NULL)
  
  #Save and clean updated IDs for use in rest of report
  save(erddap_datasets, file = file.path(data.dir, "ERDDAP_datasets.Rdata"))
  
  # Exclude stock assessment status data, which have unique structure
  erddap_datasets <- erddap_datasets %>%
    dplyr::filter(!str_detect(Dataset.ID, "assess")) 
  
  #Create SOE parent data set, filter out NAs. This queries based on 
  #data set IDs that were collected above
  SOE.data.erd <- sprintf("http://comet.nefsc.noaa.gov/erddap/tabledap/%s.csv",
                           erddap_datasets$Dataset.ID) %>% 
    purrr::map(function(x) {
      readr::read_csv(url(x))
    }) %>% 
    do.call(rbind,.) %>% 
    mutate(Value = as.numeric(Value)) %>% 
      dplyr::filter(!is.na(Value))
  
  #Convert to data.table
  SOE.data <- as.data.table(SOE.data.erd)
  
  #Save data
  save(SOE.data, file = file.path(data.dir,"SOE_data_erddap.Rdata"))
}
```

#### A note on data structures

The majority of the derived time series used in State of the Ecosystem reports are in long format. This approach was taken so that all disparate data sets could be "bound" together for ease of use in our base plotting [functions]((https://github.com/NOAA-EDAB/tech-doc/blob/master/R/BasePlot_source.R)).
