---
title: "New England shellfish bed closures due to HAB-associated toxins"
output:
  html_document: default
  word_document: default
  pdf_document: default
bibliography: NE_HAB.bib
header-includes: 
- \usepackage{float}
- \usepackage[bottom]{footmisc}
---

```{r setup,  echo = F, message=F}

#Load packages
library(knitr)
library(rmarkdown)

```

<!--The purpose of this form is to provide the methods used to collect, process, and analyze derived data prior to data submission for all EDAB reports. This information will be compiled into the State of the Ecosystem Tech Memo, AND be used to meet metadata requirements put forth by the Public Access to Research Results (PARR) directive. 

For the tech memo, we will be documenting the methods behind each indicator to the level of a peer-reviewed document.
-->
# New England Harmful Algal Bloom Indicator

## Contributor name(s)
Sean Hardison, Dave Kulis, Donald M Anderson

## Data name
New England Shellfish Bed Closures 2007-2016

### Indicator category
<!-- 1. Database pull -->
<!-- 2. Database pull with analysis -->
3. Synthesis of published information
<!-- 4. Extensive analysis, not yet published -->
<!-- 5. Published methods -->

##Methods
The New England Harmful Algal Bloom (HAB) indicator is a synthesis of shellfish bed closures related to the presence of HAB-associated toxins above threshold levels from 2007-2016. Approved detection methods were used to identify the presence of toxins associated with Amnesic Shellfish Poison (ASP), Paralytic Shellfish Poisoning (PSP), and Diarrhetic Shellfish Poisoning (DSP) by state and federal laboratories. 

####Paralytic Shellfish Poisoning
The most common cause of shellfish bed closures in New England is the presence of paralytic shellfish toxins (PSTs) produced by the dinoflagellate *Alexandrium catenella*. All New England states except Maine relied on the AOAC-approved mouse bioassay method to detect PSTs in shellfish during the 2007-2016 period reported here [@Anonymous2005].

In Maine, PST detection methods were updated in May 2014 when the state adopted the hydrophilic interaction liquid chromatography (HILIC) UPLC-MS/MS protocol [@Boundy2015] in concordance with National Shellfish Sanitation Program (NSSP) requirements.  Prior to this, the primary method used to detect PST in Maine was with the mouse bioassay. 

####Amnesic Shellfish Poisoning
The syndrome amnesic shellfish poisoning (ASP) is caused by the toxin domoic acid (DA), which is produced by several phytoplankton species belonging to the genus *Pseudo-nitzchia*. In New England, a UV-HPLC method [@Quilliam1995], which specifies a HPLC-UV protocol. 

####Diarrhetic Shellfish Poisoning
Diarrhetic Shellfish Poisoning (DSP) is rare in New England but the presence of the DSP-associated okadaic acid (OA) in mussels was confirmed in Massachusetts in 2015 (personal comm.). Preliminary testing for OA in Massachusetts utilized the commercially available Protein Phosphatase Inhibition Assay (PPIA) and these results are confirmed through LC-MS/MS when necessary [@Smienk2012; @Stutts2017]. 


### Data source(s)
<!--Please provide a text description of data sources, inlcuding primary collection methods. What equipment was used to turn signal to data? From which vessel were data collected and how? What quality control procedures were employed, if any?--> 


Data used in this indicator were drawn from the 2017 Report on the ICES-IOC Working Group on Harmful Algal Bloom Dynamics (WGHABD). The report and data are available here:

http://www.ices.dk/sites/pub/Publication%20Reports/Expert%20Group%20Report/SSGEPD/2017/01%20WGHABD%20-%20Report%20of%20the%20ICES%20-%20IOC%20Working%20Group%20on%20Harmful%20Algal%20Bloom%20Dynamics.pdf

Closure information was collated from information provided by the following organizations:
```{r source table, echo = F, include = T, results='asis'}
tabl <- "
| State     | Source Organization                                  |
|:-------------------------|:----------------------------------------------------------|
|Maine|Maine Department of Marine Resources|
|New Hampshire|New Hampshire Department of Environmental Services|
|Massachusetts|Massachusetts Division of Marine Fisheries|
|Rhode Island|Rhode Island Department of Environmental Management|
|Connecticut|Connecticut Department of Agriculture|
"
cat(tabl)
```


### Data extraction
<!--Text overview description of extraction methods. What information was extracted and how was it aggregated? Can point to other indicator extraction methods if the same.-->

Data were extracted from the original report visually and accuracy confirmed with report authors.

<!--Write SQL query here
```{sql, eval = F, echo = T}
SELECT * FROM...
```
--> 
```{r r_extract, echo = T, eval = T}

 #No code used for extraction

```


### Data analysis
<!--Text description of analysis methods, similar in structure and detail to a peer-reviewed paper methods section-->

No data analysis steps took place for this indicator.

<!--What packages or libraries did you use in your work flow?-->
```{r packages, echo=T, message=FALSE, warning=FALSE, include=T}

#SOE R packages

library(Kendall);library(data.table);library(zoo)
library(dplyr);library(nlme);library(AICcmodavg)
library(colorRamps);library(Hmisc);library(rgdal)
library(maps);library(mapdata);library(raster)
library(grid);library(stringr);library(png)
library(ncdf4);library(marmap); library(magick)

```

<!--Include accompanying R code, pseudocode, flow of scripts, and/or link to location of code used in analyses.-->

The script used to develop the figure in the SOE report is below. 

```{r NE_HAB, fig.cap="Regional HAB-related shellfish bed closures in New England (2007-2016).\\label{HAB}", echo=T,message=FALSE, warning=FALSE ,fig.align='center', fig.asp=1.1,fig.width=6,fig.pos='H',out.extra='trim={0cm 3cm 0cm 0cm},clip'}
#get map data and set constants
data.dir  <- "c:/users/sean.hardison/documents/soe_compile"
gis.dir <- 'c:/users/sean.hardison/documents/soe_compile/gis'

data.dir <- './data'
image.dir <- './images'
gis.dir <- './gis'

#SOE data
load(file.path(data.dir, "SOE_data_2018.Rdata"))

#projection
map.crs <- CRS("+proj=longlat +lat_1=35 +lat_2=45 +lat_0=40 +lon_0=-77 +x_0=0
               +y_0=0 +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0")

#coastline
coast <- readOGR(gis.dir, 'NES_LME_coast', verbose = F)
coast <- spTransform(coast,map.crs)

#define extents for cropping
e1  <- extent(-78.5, -64, 41, 45)

#crop
coast1 <- crop(coast, e1)

#Get data from SOE dataset
events <- SOE.data.2018[grepl("SP occurrence",SOE.data.2018$Var),]
events_df <- data.frame(lon = events[grepl("Lon",events$Var),]$Value,
                        lat = events[grepl("Lat",events$Var),]$Value,
                        val = events[grepl("occurrence N",events$Var),]$Value,
                        var = word(events[grepl("occurrence N", events$Var),]$Var,1))


g1 <- events_df %>% filter(val == 8)
g2 <- events_df %>% filter(val == 3.5)
g3 <- events_df %>% filter(val == 1)

#data.frame to sp object
coordinates(g1) <- ~lon+lat
g1@proj4string <- map.crs
g1 <- spTransform(g1, map.crs)

coordinates(g3) <- ~lon+lat
g3@proj4string <- map.crs
g3 <- spTransform(g3, map.crs)


#plot map and dots of different size based on category
par(mar = c(0,2.8,0,0.2))
plot(coast1, xlim = c(-71.5,-64.5),ylim = c(41,45),col = "grey",yaxs="i")
plot(g1,  add = T,cex = 6, pch = 16, col = "darkorange")
#plot(g2,  add = T,cex = 4, pch = 16, col = "purple")
plot(g3,  add = T,cex = 2, pch = 16, col = c("purple","#56B4E9"))

axis(1, at = c(-71,-69,-67,-65), labels = paste( c(-71,-69,-67,-65) * -1, 'W')
     ,col = NA, col.ticks = 1, pos = 41)
axis(2, at = c(45, 44, 43, 42, 41.05), labels = paste(c(45, 44, 43, 42, 41), 'N')
     , las = T, pos = -71.78,col = NA, col.ticks = 1)
legend(-66.2,42.5, c("PSP", "ASP", "DSP"), col = c("darkorange","purple","#56B4E9"), pch = 16,
       cex = 1.1, bty = "n", pt.cex = 2)
text(-65.65,42.495, "Category")
arrows(-71.765,41,-71.765,45,angle = 90,lwd = 2)
arrows(-64.22,41,-64.22,45,angle = 90,lwd = 2)
abline(h = 41, lwd = 2)
abline(h = 45, lwd = 2)
abline(v = -64, lwd = 2)
legend(-67.9,42.5, c("6-10", "2-5", "1"), col = c("black"), pch = 16, cex = 1.1,
       pt.cex = c(6,4,2), x.intersp = 1.45, y.intersp = 1.75, bty = "n")
text(-67.7,42.525, "2007-2016 Detections")

```


### ERDDAP data set
If data are available, adjust code to pull those data and set "eval = T" in chunk header options.
```{r erddap pull, echo = T, eval = F}
library(rerddap)

comet <- 'http://comet.nefsc.noaa.gov/erddap/'

tab_list <- ed_datasets(url = comet)
tab_list <- data.frame(Title = tab_list$Title,
                        Identifier = tab_list$Dataset.ID,
                        Index = 1:nrow(tab_list))

#Get data set ID (will need to change)
id <- as.character(tab_list$Identifier[15]) 

#Pull data
data <- tabledap(id,url = comet)

#Print metadata
metadata <- info(id, url = "http://comet.nefsc.noaa.gov/erddap/")$alldata$NC_GLOBAL
cat(metadata[metadata$attribute_name == "lineage processing steps",]$value)

```


### Further metadata
NA

#### Public availability statement
NA

#### Point of contact - Whom should be contacted?
Sean Hardison

#### Data steward (can be POC; in charge of keeping data updated)
Sean Hardison


### References
<!--List references here-->






