---
title: "Groundfish productivity"
output:
  html_document: default
  word_document: default
bibliography: productivity_tech_memo.bib
---
<!-- STOP!!!!! INSERT THE FILE NAME OF YOUR .bib BIBLIOGRAPHY ABOVE.--> 

```{r setup,  echo = F, message=F}

#Load packages
library(knitr)
library(rmarkdown)

```

## Contributor name(s)
Charles Perretti

## Data name
Groundfish productivity estimated as the ratio of small fish to large fish

### Indicator category 
Published methods


## Methods


### Data source(s) (All types)
Survey data from NEFSC trawl database (survdat).


### Data extraction 
Survdat generated by Sean Lucey.


### Data analysis
We defined size thresholds separating small and large fish for each species based on the 20th percentile of the length distribution across all years. This threshold was then used to calculate a small and large fish index (numbers below and above the threshold, respectively) each year. Although the length percentile corresponding to age-1 fish will vary with species, we use the 20th percentile as an approximation. Biomass was calculated using length–weight relationships directly from the survey data. Following Wigley et al. (2003), the length-weight relationship was modeled as 
$$\ln W = \ln a + b \ln L$$
where $W$ is weight (kg), $L$ is length (cm), and $a$ and $b$ are parameters fit via linear regression. The ratio of small fish numbers of the following year to larger fish biomass in the current year was used as the index of recruitment success. The fall and spring recruitment success anomalies were averaged to provide an annual index of recruitment success.

Further details of methods described in Perretti et al. (2017).


```{r packages, echo = T}
sessionInfo(package = NULL)
#Insert packages used here
library(ggplot2)
#library(rpart)
library(dplyr)
```


```{r analysis, echo = T, eval = T}
load("dat_spec_rec_forSOE.Rdata")
load("dat_spec_rec_epu_forSOE.Rdata")

#### Functions for plotting ####
library(ggplot2)
library(rpart)

# Adjust plot properties
adjustAxes <- 
  ggplot2::theme(axis.title   = element_text(size = 18),
                 axis.text    = element_text(size = 15),
                 plot.title   = element_text(size = 20))

ggplot <- function(...) { ggplot2::ggplot(...)  + 
    ggplot2::theme_bw() + 
    adjustAxes}


# Plot stacked bar with cpts for single var 
plot_stackbarcpts_single <- function(YEAR, var2bar,
                                     x, xlab, ylab,
                                     titl,
                                     file_suffix,
                                     leg_font_size = 10,
                                     remove_leg = FALSE,
                                     leg_ncol = 1,
                                     wcpts = TRUE,
                                     wdashed = TRUE,
                                     height = 5.5,
                                     width = 8) {
  
  dat2bar <- data.frame(YEAR, var2bar,
                        x)
  
  dat2plot <-
    dat2bar %>%
    tidyr::gather(variable, value, -YEAR, -var2bar) %>%
    dplyr::mutate(var2bar = gsub(pattern      = "_", 
                                 replacement  = " ", 
                                 x            = var2bar),
                  var2bar = gsub(pattern      = "Atl.", 
                                 replacement  = "", 
                                 x            = var2bar),
                  var2bar = gsub(pattern      = "Atl", 
                                 replacement  = "", 
                                 x            = var2bar),
                  var2bar = gsub(pattern      = "NS and combined", 
                                 replacement  = "", 
                                 x            = var2bar),
                  var2bar = gsub(pattern      = "YT", 
                                 replacement  = "Yellowtail", 
                                 x            = var2bar),
                  var2bar = gsub(pattern      = " GoM", 
                                 replacement  = " GOM", 
                                 x            = var2bar))
  
  
  p <-   
    ggplot(dat2plot,
           aes(x = YEAR)) +
    geom_bar(data = dat2plot %>% filter(value > 0),
             aes(y = value, fill = var2bar),
             stat = "identity") +
    geom_bar(data = dat2plot %>% filter(value < 0),
             aes(y = value, fill = var2bar),
             stat = "identity") +
    geom_hline(size = 0.3, aes(yintercept = 0)) +
    xlab(xlab) +
    ylab(ylab) +
    ggtitle(titl) +
    guides(fill = guide_legend(ncol = leg_ncol)) +
    theme(axis.title   = element_text(size = 16),
          axis.text    = element_text(size = 15),
          plot.title   = element_text(size = 20),
          legend.text  = element_text(size = leg_font_size),
          legend.title = element_blank())
  
  if(remove_leg) p <- p + theme(legend.position = "none")
  
  print(p)
  
  ggsave(plot = p,
         filename = "./productivity_all.eps",
         width = width,
         height = height)
}

# Plot stacked bars
plot_stackbarcpts <- function(YEAR, var2bar,
                              top, mid, bot,
                              top_lab, 
                              mid_lab, 
                              bot_lab,
                              xlab = "", 
                              ylab = "",
                              titl = "") {
  
  dat2bar <- data.frame(YEAR, var2bar,
                        top, mid, 
                        bot)
  
  dat2plot <-
    dat2bar %>%
    tidyr::gather(variable, value, -YEAR, -var2bar) %>%
    dplyr::mutate(variable = ifelse(variable == "top", 
                                    top_lab, 
                                    ifelse(variable == "mid",
                                           mid_lab, bot_lab)))
  
  
  dat2plot$variable <- 
    factor(dat2plot$variable,
           levels = c(top_lab, mid_lab, bot_lab))
  
  p <-   
    ggplot(dat2plot,
           aes(x = YEAR)) +
    geom_bar(data = dat2plot %>% filter(value > 0),
             aes(y = value, fill = var2bar),
             stat = "identity") +
    geom_bar(data = dat2plot %>% filter(value < 0),
             aes(y = value, fill = var2bar),
             stat = "identity") +
    facet_wrap(~ variable, ncol = 1) +
    geom_hline(size = 0.3, aes(yintercept = 0)) +
    xlab(xlab) +
    ylab(ylab) +
    ggtitle(titl) +
    guides(fill = guide_legend(ncol = 1)) +
    theme(axis.title   = element_text(size = 16),
          axis.text    = element_text(size = 15),
          plot.title   = element_text(size = 20),
          strip.text   = element_text(size = 15),
          legend.title = element_blank())
  
  print(p)
  
  ggsave(plot = p,
         filename = "./productivity_epu.eps",
         width = 7,
         height = 9)
}

# Recruit per spawner (all stocks in one figure)
plot_stackbarcpts_single(YEAR = dat_spec_rec_forSOE$Time,
                         var2bar = dat_spec_rec_forSOE$Var,
                         x = dat_spec_rec_forSOE$Value,
                         titl = "",
                         xlab = "",
                         ylab = "Small fish per large fish biomass (anomaly)",
                         height = 5.5,
                         width = 9)


# Recruit per spawner (by EPU)
dat2plot <-
  dat_spec_rec_epu_forSOE %>%
  tidyr::spread(Region, Value)


plot_stackbarcpts(YEAR = dat2plot$Time,
                  var2bar = dat2plot$Var,
                  top = dat2plot$GOM, 
                  mid = dat2plot$GB,
                  bot = dat2plot$MAB,
                  top_lab = "Gulf of Maine",
                  mid_lab = "Georges Bank",
                  bot_lab = "Mid. Atlantic Bight",
                  xlab = "",
                  ylab = "Small fish per Large fish biomass (anomaly)")

```



### Further metadata

#### Public availability statement
Source data available to public.


#### Point of contact - Whom should be contacted?
Charles Perretti

#### Data steward (can be POC; in charge of keeping data updated)
Charles Perretti

#### References
<!-- References must be submitted in BibTeX format. When compiling this document, they will appear in this section. There is no need to write them out here. Don't forget to include the name of your .bib file in the YAML header (fill in XYZ.bib in YAML header)!!!!!--> 






