# Fish Productivity Indicator
 

```{r setup,  echo = F, message=F}

#Load packages
library(knitr)
library(rmarkdown)

```
**Description**: Groundfish productivity estimated as the ratio of small fish to large fish

**Indicator category**: Published methods

**Contributor(s)**: Charles Perretti
  
**Data steward**: Charles Perretti, <charles.perretti@noaa.gov>
  
**Point of contact**: Charles Perretti, <charles.perretti@noaa.gov>
  
**Public availability statement**: Source data available to public.



## Methods


### Data sources
Survey data from NEFSC trawl database. These data in their derived form are available through [Survdat](#survdat).


### Data extraction 
Data were extracted from [Survdat](#survdat).


### Data analysis
We defined size thresholds separating small and large fish for each species based on the 20th percentile of the length distribution across all years. This threshold was then used to calculate a small and large fish index (numbers below and above the threshold, respectively) each year. Although the length percentile corresponding to age-1 fish will vary with species, we use the 20th percentile as an approximation. Biomass was calculated using length–weight relationships directly from the survey data. Following @wigley_length-weight_2003, the length-weight relationship was modeled as 
$$\ln W = \ln a + b \ln L$$
where $W$ is weight (kg), $L$ is length (cm), and $a$ and $b$ are parameters fit via linear regression. The ratio of small fish numbers of the following year to larger fish biomass in the current year was used as the index of recruitment success. The fall and spring recruitment success anomalies were averaged to provide an annual index of recruitment success.

Further details of methods described in @perretti_regime_2017.

### Plotting

```{r analysis, echo = T, eval = T, message=F, warning=F, fig.cap = "Groundfish productivity across all stocks in the Northeast Large Marine Ecosystem."}

data.dir  <- here::here('data')

load(file.path(data.dir,"dat_spec_rec_forSOE.Rdata"))
load(file.path(data.dir,"dat_spec_rec_epu_forSOE.Rdata"))

#### Functions for plotting ####
library(ggplot2)
library(rpart)
library(dplyr)

# Adjust plot properties
adjustAxes <- 
  ggplot2::theme(axis.title   = element_text(size = 18),
                 axis.text    = element_text(size = 15),
                 plot.title   = element_text(size = 20))

ggplot <- function(...) { ggplot2::ggplot(...)  + 
    ggplot2::theme_bw() + 
    adjustAxes}


# Plot stacked bar with cpts for single var 
plot_stackbarcpts_single <- function(YEAR, var2bar,
                                     x, xlab, ylab,
                                     titl,
                                     file_suffix,
                                     leg_font_size = 10,
                                     remove_leg = FALSE,
                                     leg_ncol = 1,
                                     wcpts = TRUE,
                                     wdashed = TRUE,
                                     height = 5.5,
                                     width = 8) {
  
  dat2bar <- data.frame(YEAR, var2bar,
                        x)
  
  dat2plot <-
    dat2bar %>%
    tidyr::gather(variable, value, -YEAR, -var2bar) %>%
    dplyr::mutate(var2bar = gsub(pattern      = "_", 
                                 replacement  = " ", 
                                 x            = var2bar),
                  var2bar = gsub(pattern      = "Atl.", 
                                 replacement  = "", 
                                 x            = var2bar),
                  var2bar = gsub(pattern      = "Atl", 
                                 replacement  = "", 
                                 x            = var2bar),
                  var2bar = gsub(pattern      = "NS and combined", 
                                 replacement  = "", 
                                 x            = var2bar),
                  var2bar = gsub(pattern      = "YT", 
                                 replacement  = "Yellowtail", 
                                 x            = var2bar),
                  var2bar = gsub(pattern      = " GoM", 
                                 replacement  = " GOM", 
                                 x            = var2bar))
  
  
  p <-   
    ggplot(dat2plot,
           aes(x = YEAR)) +
    geom_bar(data = dat2plot %>% filter(value > 0),
             aes(y = value, fill = var2bar),
             stat = "identity") +
    geom_bar(data = dat2plot %>% filter(value < 0),
             aes(y = value, fill = var2bar),
             stat = "identity") +
    geom_hline(size = 0.3, aes(yintercept = 0)) +
    xlab(xlab) +
    ylab(ylab) +
    ggtitle(titl) +
    guides(fill = guide_legend(ncol = leg_ncol)) +
    theme(axis.title   = element_text(size = 16),
          axis.text    = element_text(size = 15),
          plot.title   = element_text(size = 20),
          legend.text  = element_text(size = leg_font_size),
          legend.title = element_blank())
  
  if(remove_leg) p <- p + theme(legend.position = "none")
  
  print(p)
  
#  ggsave(plot = p,
#         filename = "./productivity_all.eps",
#         width = width,
#         height = height)
}

# Plot stacked bars
plot_stackbarcpts <- function(YEAR, var2bar,
                              top, mid, bot,
                              top_lab, 
                              mid_lab, 
                              bot_lab,
                              xlab = "", 
                              ylab = "",
                              titl = "") {
  
  dat2bar <- data.frame(YEAR, var2bar,
                        top, mid, 
                        bot)
  
  dat2plot <-
    dat2bar %>%
    tidyr::gather(variable, value, -YEAR, -var2bar) %>%
    dplyr::mutate(variable = ifelse(variable == "top", 
                                    top_lab, 
                                    ifelse(variable == "mid",
                                           mid_lab, bot_lab)))
  
  
  dat2plot$variable <- 
    factor(dat2plot$variable,
           levels = c(top_lab, mid_lab, bot_lab))
  
  p <-   
    ggplot(dat2plot,
           aes(x = YEAR)) +
    geom_bar(data = dat2plot %>% filter(value > 0),
             aes(y = value, fill = var2bar),
             stat = "identity") +
    geom_bar(data = dat2plot %>% filter(value < 0),
             aes(y = value, fill = var2bar),
             stat = "identity") +
    facet_wrap(~ variable, ncol = 1) +
    geom_hline(size = 0.3, aes(yintercept = 0)) +
    xlab(xlab) +
    ylab(ylab) +
    ggtitle(titl) +
    guides(fill = guide_legend(ncol = 1)) +
    theme(axis.title   = element_text(size = 16),
          axis.text    = element_text(size = 15),
          plot.title   = element_text(size = 20),
          strip.text   = element_text(size = 15),
          legend.title = element_blank())
  
  print(p)
}

# Recruit per spawner (all stocks in one figure)
plot_stackbarcpts_single(YEAR = dat_spec_rec_forSOE$Time,
                         var2bar = dat_spec_rec_forSOE$Var,
                         x = dat_spec_rec_forSOE$Value,
                         titl = "",
                         xlab = "",
                         ylab = "Small fish per large fish biomass (anomaly)",
                         height = 7,
                         width = 9)

```

```{r plot-all, fig.height=8, fig.cap = "Groundfish productivity visualized by EPU."}


# Recruit per spawner (by EPU)
dat2plot <-
  dat_spec_rec_epu_forSOE %>%
  tidyr::spread(Region, Value)


plot_stackbarcpts(YEAR = dat2plot$Time,
                  var2bar = dat2plot$Var,
                  top = dat2plot$GOM, 
                  mid = dat2plot$GB,
                  bot = dat2plot$MAB,
                  top_lab = "Gulf of Maine",
                  mid_lab = "Georges Bank",
                  bot_lab = "Mid. Atlantic Bight",
                  xlab = "",
                  ylab = "Small fish per Large fish biomass (anomaly)")

```

