# Seasonal SST Anomalies


```{r setup, echo = F, message=F}

#Load packages
library(knitr)
library(rmarkdown)

```
**Description**: Seasonal SST Anomalies

**Indicator category**: Database pull with analysis

**Contributor(s)**: Sean Hardison, Vincent Saba
  
**Data steward**: Sean Hardison, <sean.hardison@noaa.gov>
  
**Point of contact**: Sean Hardison, <sean.hardison@noaa.gov>
  
**Public availability statement**: Source data are available [here](https://www.esrl.noaa.gov/psd/data/gridded/data.noaa.oisst.v2.highres.html).




## Methods
### Data sources
Data for seasonal sea surface tempature anomalies were derived from the NOAA optimum interpolation sea surface temperature high resolution data set ([NOAA OISST V2](https://www.esrl.noaa.gov/psd/data/gridded/data.noaa.oisst.v2.highres.html)) provided by NOAA/OAR/ESRL PSD, Boulder, CO. The data extend from 1981 to present, and provide a 0.25&deg; x 0.25&deg; global grid of SST measurements [@Reynolds2007].  


### Data extraction 
Individual files containing daily mean SST data for each year during the period of 1981-present were downloaded from the [OI SST V5 site](http://www1.ncdc.noaa.gov/pub/data/cmb/ersst/v5/netcdf/). Yearly data provided as layered rasters were masked according to the extent of Northeast US Continental Shelf (See [Seasonal_anom_crop.nc](https://github.com/NOAA-EDAB/tech-doc/tree/master/gis)). Data were split into three month seasons for analysis (Winter = Jan, Feb, Mar; Spring = Apr, May, Jun; Summer = July, August, September; Fall = Oct, Nov, Dec).  


R code used in extraction and processing:
```{r extraction, echo = T, eval = F}
#libraries
library(ncdf4);library(dplyr)
library(readr);library(tidyr)
library(sp);library(rgdal)
library(raster);library(stringr)

fall <- stack()
winter <- stack()
spring <- stack()
summer <- stack()
sub_region <- raster("Seasonal_anom_crop.nc") #get an old file to crop with

for (dir. in 1){
  #Loop through directories
  setwd(paste0('c:/users/sean.hardison/documents/sst_data/',dir.))
  print(getwd())
  for (f in 1:length(list.files())){
    
    if (!str_detect(list.files()[f],".nc")){
      print(paste(list.files()[f],"is not a raster")) #Based on file type
      next
    }
    
    for (s in c("winter","spring","summer","fall")){
  
      y <- as.numeric(str_extract(list.files()[f],"[0-9]+")) #Pull out file year for identifying leap years
      print(paste('Year:',y,"Season:",s))
      leap_year <- seq(1984,2016,4)
      
      #Pick out leap years
      if (y %in% leap_year){
          leap <- T
      } else {
        leap <- F
      }
      if (s == "winter")
        if (!leap){
          season <- 1:90 
        } else {
          season <- 1:91
        }
      else if (s == "spring"){
        if (!leap){
          season <- 91:181 
        } else {
          season <- 92:182
        } 
      }
      else if (s == "summer"){
        if (!leap){
          season <- 182:273
        } else {
          season <- 183:274
        } 
      }
      else if (s == "fall"){
        if (!leap){
          season <- 274:365
        } else {
          season <- 275:366
        } 
      }
      
      if(leap){
        print(paste(s,y,"leap"))
      }
      
        
        for (b in season){
          r <- raster(list.files()[f],b) #Import raster bands within a given season

          if (b == season[1]){
            print(list.files()[f])
            print(length(season))
            print(r@file@nbands) #Print file name
          }
          
          #Rotate and crop
          r <- rotate(r) 
          r <- crop(r, extent(sub_region))
          #print(r@file@nbands)
          if (b == season[50]){
            print("Over halfway there")
          }
          
          #Add bands to raster stack for each season. Seaonal bands for each year are bound together.
          assign(s,stack(get(s),r))
          
        }
      print(paste("Done stacking",nlayers(get(s)),"days"))
    } 
  }
}


```

### Data analysis
We calculated the long-term mean (LTM) for each season-specific stack of rasters for the period of 1982-2012, and then subtracted the (LTM) from 2017 daily mean SST values to find the 2017 SST anomaly. 

```{r anomaly calculation, echo = T, eval = F}
fall_1 <- stack("fall_1")
fall_2 <- stack("fall_2")
fall_3 <- stack("fall_3")


spring_1 <- stack("spring_1")
spring_2 <- stack("spring_2")
spring_3 <- stack("spring_3")


winter_1 <- stack("winter_1")
winter_2 <- stack("winter_2")
winter_3 <- stack("winter_3")


summer_1 <- stack("summer_1")
summer_2 <- stack("summer_2")
summer_3 <- stack("summer_3")

winter <- stack(winter_1,winter_2,winter_3)
fall <- stack(fall_1,fall_2,fall_3)
spring <- stack(spring_1,spring_2,spring_3)
summer <- stack(summer_1,summer_2,summer_3)

winter_mean <- stackApply(winter, indices = rep(1,nlayers(winter)),mean)
spring_mean <- stackApply(spring, indices = rep(1,nlayers(spring)),mean)
summer_mean <- stackApply(summer, indices = rep(1,nlayers(summer)),mean)
fall_mean <- stackApply(fall, indices = rep(1,nlayers(fall)),mean)

rm(winter, spring, summer, fall)

#Get 2017 daily mean data
#setwd("z:/shardison/neiea")
daily_mean_2017 <- stack("sst.day.mean.2017.nc")



daily_mean_2017 <- rotate(daily_mean_2017) #rotate
daily_mean_2017 <- crop(daily_mean_2017, extent(sub_region)) #crop

#Find daily mean across seasons
seasonal_sst_2017 <- stackApply(daily_mean_2017, indices = c(rep(1,90),rep(2,91),
                                                             rep(3, 92), rep(4, 92)), mean)

#Get seasonal anomaly
winter_anom <- seasonal_sst_2017[[1]] - winter_mean
spring_anom <- seasonal_sst_2017[[2]] - spring_mean
summer_anom <- seasonal_sst_2017[[3]] - summer_mean
fall_anom <- seasonal_sst_2017[[4]] - fall_mean
```

### Plotting
```{r sst-anomaly-maps, fig.cap="Sea surface temperature anomalies in winter (A), spring (B), summer (C), and fall (D) 2017." , fig.width = 8,fig.height = 5.5, fig.show='hold',fig.pos='H', echo=T, message=F, warning=F}


# Relative working directory
r.dir <- "R"

#Source constants for GIS code
source(file.path(r.dir,"GIS_source.R"))

#SST map function
SST_plot <- function(raster_data, text,pos = "left"){
  
  #adjusts par for legend placement 
  if (pos == "right"){
    par(mar = c(4.5,0,3,10), mex = .3,  mgp = c(4, .35, 0))
  } else{
     par(mar = c(4.1,5,3,6), mex = .3,  mgp = c(4, .35, 0))
  }
  
  #read .nc and convert to raster
  d <- raster(file.path(gis.dir,raster_data))
  crs(d) <- map.crs
  #crop data
  d <- crop(d, e)
  
  #get zmax (~max SST anomaly)
  z.max=7

  #plot raster
  plot(d, col=colors(n=color_levels),
       breaks=seq(-z.max,z.max,length.out=color_levels+1),
       legend.width = 1.5,bty ="L",axes=F,
       interpolate = T, las = 1,box = FALSE,legend = FALSE, asp = 1)
  #plot contour
  contour(bathy,drawlabels = F,nlevels = 5,add=T)
  #plot coastline
  plot(coast, col = 'grey',add = T , lty=0)
  #axes

  
  if (pos == "left"){
    axis(1, at = seq(-77,-65,2), labels = seq(-77,-65,2) , cex.axis = 0.7)
    axis(2, cex.axis = 0.7)
    abline(h = 34.95, lty = 1, lwd = 2)
    abline(v =-77.2, lty = 1, lwd = 2)
  } else {
    axis(1, at = seq(-77,-65,2), labels = seq(-77,-65,2) ,
         cex.axis = 0.7, pos = 34.9)
    axis(2, cex.axis = 0.7, pos = -77.225)
    abline(h = 34.9, lty = 1, lwd = 2)
    abline(v =-77.225, lty = 1, lwd = 2)

  }

  #labels
  text(-76,45,text,cex = 1.5)

  
}

#plot maps
par(oma=c(0,0,0,0), mfrow = c(2,2))
SST_plot("Winter_2017_anomaly_1982_2012.nc","A")
SST_plot("Spring_2017_anomaly_1982_2012.nc","B",pos = "right")
SST_plot("Summer_2017_anomaly_1982_2012.nc","C")
SST_plot("Fall_2017_anomaly_1982_2012.nc","D",pos = "right") 

#plot only legend
z.max=7

#set new par
par(mfrow=c(1, 1), mar=c(2, 5, 2, 0), new=FALSE)

#any data
d <- raster(matrix(runif(100), ncol=10))

#legend plot
plot(d, col=colors(n=color_levels),
     breaks=seq(-z.max,z.max,length.out=color_levels+1),
     legend.shrink = 1,
     legend.width = 1,legend.only = T,
     axes=F, interpolate = T, las = 1,box = FALSE,
     axis.args=list(at=seq(-z.max, z.max, length.out = 5),
                    labels=seq(-7,7,length.out = 5), 
                    cex.axis=0.8),
     legend.args=list(text=expression(paste('Temperature Anomaly (',degree,"C)")),
                      side=4, font=2, line=2, cex=1))

```

**Note**: The derived data sets used to create the above figures are available [here](https://github.com/NOAA-EDAB/tech-doc/tree/master/gis).
