# Single Species Status Indicator {#stockstatus}

```{r setup,  echo = F, message=F}

#Load packages
library(knitr)
library(rmarkdown)

```


**Description**: Most recent stock assessment model results summarized for each assessed species.

**Indicator category**: Synthesis of published information

**Contributor(s)**: Sarah Gaichas, based on code and spreadsheets originally provided by Chris Legault

**Data steward**: Sarah Gaichas <sarah.gaichas@noaa.gov>

**Point of contact**: Sarah Gaichas <sarah.gaichas@noaa.gov>

**Public availability statement**: All stock assessment results are publicly available (see Data Sources). Summarized data are available [here](http://comet.nefsc.noaa.gov/erddap/tabledap/assess_soe_v1.htmlTable?No,Entity_Name,Science_Center,Assessment_Year,Last_Data_Year,Assessment_Level,Citation,Comments,Best_F,F_Year,Flimit,Fmsy,F_Flimit,F_Fmsy,Best_B,B_Year,B_Blimit,B_Bmsy,Stock_Level_Relative_to_Bmsy,Bmsy,Blim).

## Methods

### Data sources
<!--Please provide a text description of data sources, inlcuding primary collection methods. What equipment was used to turn signal to data? From which vessel were data collected and how? What quality control procedures were employed, if any?--> 

"Data" used for this indicator are the outputs of stock assessment models and review processes, including reference points (proxies for fishing mortality limits and stock biomass targets and limits), and the current fishing mortality rate and biomass of each stock. The spreadsheet summarizes most recent stock assessment updates for each species, which are available on the Northeast Fisheries Science Center (NEFSC) website at:
https://www.nefsc.noaa.gov/saw/reports.html  
https://www.nefsc.noaa.gov/publications/crd/crd1717/

Additional assessments are reported directly to the New England Fishery Management Council (NEFMC):
http://s3.amazonaws.com/nefmc.org/Document-2-SAFE-Report-for-Fishing-Year-2016.pdf  
http://s3.amazonaws.com/nefmc.org/4_NEFSC_SkateMemo_July_2017.pdf



### Data extraction
<!--Text overview description of extraction methods. What information was extracted and how was it aggregated? Can point to other indicator extraction methods if the same.-->

Each assessment document was searched to find the following information (often but not always summarized under a term of reference to determine stock status in the executive summary):

*    **Bcur**: current year biomass, (most often spawning stock biomass (SSB) or whatever units the reference points are in)

*    **Fcur**: current year fishing mortality, F

*    **Bref**: biomass reference point, a proxy of Bmsy (the target)

*    **Fref**: fishing mortality reference point, a proxy of Fmsy

<!--Write SQL query here
```{sql, eval = F, echo = T}
SELECT * FROM...
```
--> 

### Data analysis
<!--Text description of analysis methods, similar in structure and detail to a peer-reviewed paper methods section-->

For each assessed species, Bcur is divided by Bref and Fcur is divided by Fref. They are then plotted for each species on an x-y plot, with Bcur/Bref on the x axis, and Fcur/Fref on the y axis. 

<!--What packages or libraries did you use in your work flow?-->
```{r packages, eval=F, echo = F}
#all packages in the SOE file that ran my code snippet, probably only need data.table
library(Kendall);library(data.table);library(zoo)
library(dplyr);library(nlme);library(AICcmodavg)
library(colorRamps);library(Hmisc);library(rgdal)
library(maps);library(mapdata);library(raster)
library(grid);library(stringr);library(png)
library(ncdf4);library(marmap); library(magick)


```

<!--Include accompanying R code, pseudocode, flow of scripts, and/or link to location of code used in analyses.-->

### Plotting

The script used to develop the figure in the SOE is below, with an example figure. Different lines are commented out of the script to produce the Mid- Atlantic or New England figures. The positioning table used to make the plot (the data frame named "decoder" below) is available [here](http://comet.nefsc.noaa.gov/erddap/tabledap/assess_support_soe_v1.htmlTable?Entity_Name,Count_of_No,Min_of_No2,Code,Council,my_pos,my_pos2,my_pos3).

```{r analysis, echo = T, eval = T, fig.cap="Summary of single species status for NEFMC stocks \\label{KOBE}", fig.height=8, fig.width=11, fig.align='center', fig.show='hold', message=F, warning=F}

# Relative working directories
data.dir  <- here::here('data')
image.dir <- here::here('images')
r.dir <- here::here('R')

#Load current SOE ERDDAP data set IDs
load(file.path(data.dir, "ERDDAP_datasets.Rdata"))

#Source get_erddap function to query data
source(file.path(r.dir,"get_erddap.R"))

#Data queries
dat <- get_erddap(df = erddap_datasets,
                  id = "assess_soe_v1")

decoder <- get_erddap(erddap_datasets,
                      id = "assess_support_soe_v1")

#Plotting code
stocks <- unique(dat$Entity_Name)
n.stocks <- length(stocks)

# set up the headers for most.recent structure - will contain oldest Assessment.Year
most.recent <- dat[1,]
most.recent <- most.recent[-1,]

for (i in 1:n.stocks){
  temp <- dat[dat$Entity_Name == decoder$Entity_Name[i],]
  most.recent <- rbind(most.recent,temp[temp$Assessment_Year == max(temp$Assessment_Year),])
}
#cbind(n.stocks,length(most.recent[,1]))

# get the max of F.Flimit and F.Fmsy and the max of B.Blimit and B.Bmsy
Frat <- most.recent$F_Fmsy
Brat <- most.recent$B_Bmsy
#Frat[6] <- most.recent$F.Flimit[6]  # Ocean Quahog use Flimit because no Fmsy
#Frat[18] <- 0.88  # hardwire average of the two GOM models for GOM cod
#Brat[18] <- 0.155 # hardwire average of the two GOM models for GOM cod
#model averages have been calculated and included in F/Fmsy column in 2017assess.csv
#cbind(most.recent$Entity.Name,most.recent$F.Flimit,most.recent$F.Fmsy)

# figure out appropriate range for axes WARNING HARD CODED
#max(Frat,na.rm=T)
#max(Brat,na.rm=T)
x.r <- c(0,7)
y.r <- c(0,3.5)

# set some colors
my.col <- rep(NA,n.stocks)
MA.col <- "blue"
NE.col <- "blue"
BO.col <- "purple"
for (i in 1:n.stocks){
  if(decoder$Council[i] == "MAFMC") my.col[i] <- MA.col
  if(decoder$Council[i] == "NEFMC") my.col[i] <- NE.col
  if(decoder$Council[i] == "Both")  my.col[i] <- BO.col
}

# create new matrix for plotting
xx <- as.data.frame(cbind(decoder[,2:8],Frat,Brat,my.col))
newlist <- as.factor(decoder$Code)
others <- as.factor("7 Skates")
sppcodes <- unlist(list(newlist, others))
MA.unknown <- sppcodes[c(1,3,4,47,48)] #mackerel, Loligo, Illex, monkfish unknown
#MA.unknown <- decoder$Code[c(1,3,4,8)] #BSB known
#NE.unknown <- c(decoder$Code[c(14,23,33)],"7 Skates")
NE.overfished <- sppcodes[c(17,19,35)] # GB cod, halibut, witch, thorny skate
NE.unknown <- sppcodes[c(14,23,33,47,48)] #Red deepsea crab,  Offshore hake, GOM winter flounder, GB YT,2 monkfish, 7 skates unknown
my.col[50] <- NE.col


# break out the Councils
MA <- xx[xx$Council == "MAFMC",]
NE <- xx[xx$Council == "NEFMC",]
BO <- xx[xx$Council == "Both",]
MABO<-rbind(MA, BO)
NEBO<-rbind(NE, BO)

#if(saveplots) png(file = "MAFMC_joint_stocks_2017.png",  units="in", width = 6, height = 6, res=1200)
# plot(MABO$Brat,MABO$Frat,xlim=x.r,ylim=y.r,xlab="B/Bmsy",ylab="F/Fmsy",pch=16,col=my.col[c(1:11,47:49)])
#   abline(v=0.5,lty=2)
#   abline(v=1,lty=3)
#   abline(h=1,lty=2)
#   title(expression("MAFMC " * phantom("and Joint Stocks")), col.main="blue")
#   title(expression(phantom("MAFMC ") * "and " * phantom("Joint Stocks")), col.main="black")
#   title(expression(phantom("MAFMC and ") * "Joint" * phantom(" Stocks")), col.main="purple")
#   title(expression(phantom("MAFMC and Joint") * " Stocks"), col.main="black")  
#   legend('topright',legend=MA.unknown,pch=NA,text.col=my.col[c(1,3,4,47,48)], title="Unknown Status", title.col = "black")
#   text(MABO$Brat,MABO$Frat,labels=MABO$Code,pos=MABO$my.pos3,col=my.col[c(1:11,47:49)],cex=0.8)
#if(saveplots) savePlot("MAFMC_Joint_stocks.png", type='png')
#if(saveplots) dev.off()

#if(saveplots) pdf(file = "NEFMC_joint_stocks.pdf",  width = 6, height = 6)
#if(saveplots) png(file = "NEFMC_joint_stocks.png",  units="in", width = 6, height = 9, res=1200)
plot(NEBO$Brat,NEBO$Frat,xlim=x.r,ylim=y.r,xlab="B/Bmsy",ylab="F/Fmsy",pch=16,col=my.col[12:49])
   abline(v=0.5,lty=2)
   abline(v=1,lty=3)
   abline(h=1,lty=2)
   title(expression("NEFMC " * phantom("and Joint Stocks")), col.main="blue")
   title(expression(phantom("NEFMC ") * "and " * phantom("Joint Stocks")), col.main="black")
   title(expression(phantom("NEFMC and ") * "Joint" * phantom(" Stocks")), col.main="purple")
   title(expression(phantom("NEFMC and Joint") * " Stocks"), col.main="black")
   legend('topright',legend=NE.overfished,pch=NA,text.col=my.col[c(17,19,35)], title="Overfished, F \nStatus Unknown",title.col = "black", bty="n", text.width=strwidth("Offshore Hake"), inset=c(0,.05))
   legend('right',legend=NE.unknown,pch=NA,text.col=my.col[c(14,23,33,47,48)],title="Unknown Status", title.col = "black",bty="n")
   text(NEBO$Brat,NEBO$Frat,labels=NEBO$Code,pos=NEBO$my_pos3,col=my.col[12:49],cex=0.8)
#if(saveplots) savePlot("NEFMC_Joint_stocks.png", type='png')
#if(saveplots) dev.off()

```

