# Trend Analysis

```{r setup,  echo = F, message=F}

#Load packages
library(knitr)
library(rmarkdown)

```
**Description**: Time series trend analysis

**Indicator category**: 

<!-- 1. Database pull -->
<!-- 2. Database pull with analysis -->
<!-- 3. Synthesis of published information -->
<!-- 4. Extensive analysis, not yet published -->
<!-- 5. Published methods -->
Contributor(s): Sean Hardison, Charles Perretti, Geret DePiper

Data steward: NA

Point of contact: Sean Hardison, sean.hardison@noaa.gov

Public availability statement: NA


## Methods
In some cases, time series indicators in the State of the Ecosystem reports are presented with trend lines, although not all time series were selected for trend analysis. As has been shown elsewhere [see @Nicholson2004; @Wagner2013; @VonStorch1999a], power of statistical tests for trend is hampered by low sample size and autocorrelated observations, which are common attributes of SOE indicators. 

In a preliminary simulation study, we explored the effect of time series length and autocorrelation strength on statistical power of three trend detection methods. Overall, when sample size was low (N = 10) there was a high rate of false trend detection, and similarly, a low rate of true trend detection. Both of these forms of error were further amplified by autocorrelation in the trend residuals. Based on these findings, we selected a minimum series length of N = 30 for indicator time series before assessing trend. We also chose to use a GLS model selection (GLS-MS) approach to assess for trend in State of the Ecosystem reports, as this approach performed best overall. GLS-MS also allows for both linear and quadratic model fits and quantification of uncertainty in the trend estimate. 

The model selection procedure for the GLS approach fits four models to each time series and selects the best fitting model using AICc. The models are, 1) linear trend with uncorrelated residuals, 2) linear trend with correlated residuals, 3) quadratic trend with uncorrelated residuals, and 4) quadratic trend with correlated residuals. I.e., the models are of the form

$$ Y_t = \alpha_0 + \alpha_1X_t + \alpha_2X_t^2 + \epsilon_t$$
$$\epsilon_t = \rho\epsilon_{t-1} + \omega_t$$

$$w_t \sim N(0, \sigma^2)$$

Where $Y_t$ is the observation in time $t$, $X_t$ is the time index, $\epsilon_t$ is the residual in time $t$, and $\omega_t$ is a normally distributed random variable. Setting $\alpha_2 = 0$ yields the linear trend model, and $\rho = 0$ yields the uncorrelated residuals model.

The best fit model was tested against the null hypothesis of no trend through a likelihood ratio test (p < 0.05). All models were fit using the R package *nlme* [@Pinheiro2017] and AICc was calculated using the R package *AICcmodavg* [@Mazerolle2017a]. In SOE time series figures, significant positive trends were colored orange, and negative trends purple. 

### Data source(s)
NA

### Data extraction
NA

### Data analysis


```{r packages, echo=T, message=FALSE, warning=FALSE, include=T}

#R packages
library(dplyr)
library(nlme)
library(AICcmodavg)
library(data.table)
```

<!--Include accompanying R code, pseudocode, flow of scripts, and/or link to location of code used in analyses.-->
```{r analysis, echo = T, message= F, warning=F, include=T}

data.dir <- "./data"
load(file.path(data.dir, "SOE_data_2018.Rdata"))

#--------------------------------GLS Model Selection-----------------------------#
fit_lm <- function(dat) {

  constant_norm <-
    nlme::gls(series ~ 1, 
              data = dat)
  
  constant_ar1 <-
    try(nlme::gls(series ~ 1,
                  data = dat,
                  correlation = nlme::corAR1(form = ~time)))
  if (class(constant_ar1) == "try-error"){
    return(best_lm <- data.frame(model = NA,
                                 aicc  = NA,
                                 coefs..Intercept = NA,
                                 coefs.time = NA,
                                 coefs.time2 = NA,
                                 pval = NA)) 
  } 
  
  
  
  # Linear model with normal error
  linear_norm <- 
    nlme::gls(series ~ time, 
              data = dat)
  
  # Linear model with AR1 error
  linear_ar1 <- 
    try(nlme::gls(series ~ time, 
                  data = dat,
                  correlation = nlme::corAR1(form = ~time)))
  if (class(linear_ar1) == "try-error"){
    return(best_lm <- data.frame(model = NA,
                                 aicc  = NA,
                                 coefs..Intercept = NA,
                                 coefs.time = NA,
                                 coefs.time2 = NA,
                                 pval = NA))
    
  }
  
  # Polynomial model with normal error
  dat$time2 <- dat$time^2
  poly_norm <- 
    nlme::gls(series ~ time + time2, 
              data = dat)
  
  # Polynomial model with AR1 error
  poly_ar1 <- 
    try(nlme::gls(series ~ time + time2, 
                  data = dat,
                  correlation = nlme::corAR1(form = ~time)))
  if (class(poly_ar1) == "try-error"){
    return(best_lm <- data.frame(model = NA,
                                 aicc  = NA,
                                 coefs..Intercept = NA,
                                 coefs.time = NA,
                                 coefs.time2 = NA,
                                 pval = NA))
    
  }
  
  # Calculate AICs for all models
  df_aicc <-
    data.frame(model = c("poly_norm",
                         "poly_ar1",
                         "linear_norm",
                         "linear_ar1"),
               aicc  = c(AICc(poly_norm),
                         AICc(poly_ar1),
                         AICc(linear_norm),
                         AICc(linear_ar1)),
               coefs = rbind(coef(poly_norm),
                             coef(poly_ar1),
                             c(coef(linear_norm), NA),
                             c(coef(linear_ar1),  NA)),
               # Calculate overall signifiance (need to use
               # ML not REML for this)
               pval = c(anova(update(constant_norm, method = "ML"),
                              update(poly_norm, method = "ML"))$`p-value`[2],
                        anova(update(constant_ar1, method = "ML"),
                              update(poly_ar1, method = "ML"))$`p-value`[2],
                        anova(update(constant_norm, method = "ML"),
                              update(linear_norm, method = "ML"))$`p-value`[2],
                        anova(update(constant_ar1, method = "ML"),
                              update(linear_ar1, method = "ML"))$`p-value`[2]))
  
  best_lm <-
    df_aicc %>%
    dplyr::filter(aicc == min(aicc))
  
  
  if (best_lm$model == "poly_norm") {
    model <- poly_norm
  } else if (best_lm$model == "poly_ar1") {
    model <- poly_ar1
  } else if (best_lm$model == "linear_norm") {
    model <- linear_norm
  } else if (best_lm$model == "linear_ar1") {
    model <- linear_ar1
  }
  
  return(list(p = best_lm$pval,
              model = model))
}

#-------------------------------------Plotting code------------------------------------#
soe.plot <- function(data, x.var, y.var, x.label = '', y.label = '', tol = 0.1,
                     x.start = NA, x.end = NA, end.start = 2008, bg.col = background, mean_line = T,
                     end.col = recent, stacked = NA, x.line = 2.5, y.line = 3.5, scale.axis = 1,
                     rel.y.num = 1.5, rel.y.text = 1.5, suppressAxis = FALSE,status  = F,anomaly = F,
                     endshade = TRUE, full.trend = TRUE, point.cex = 1.5, lwd = 2, ymax = TRUE,ymin = TRUE,
                     y.upper = y.upper, y.lower = y.lower, extra = FALSE, x.var2 = x.var2, y.var2 = y.var2,
                     line.forward = FALSE, mean_line.2 = T, cex.stacked = 1, website = T) {
  

  #Select Data
  x <- data[Var == y.var, ]
  x <- x[order(x[, get(x.var)]), ]
  setnames(x, x.var, 'X')
  
  #Set common time step if necessary
  if(is.na(x.start)) x.start <- min(x[, X])
  if(is.na(x.end))   x.end   <- max(x[, X])
  x <- x[X >= x.start, ]
  
  #Set up plot parameters
  if (ymax == TRUE){
    y.max <- max(x[, Value]) + tol * max(x[, Value])
  } else {
    y.max <- as.numeric(y.upper)
  }
  
  if (ymin == TRUE){
    y.min <- min(x[, Value]) - tol * abs(min(x[, Value]))
  } else if (ymin == FALSE){
    y.min <- as.numeric(y.lower)
  }
  
  y.mean <- mean(x[, Value])
  y.sd <- sd(x[, Value])
  
  #Plot blank plot
  plot(x[X >= x.start, list(X, Var)], xlim = c(x.start, x.end),
       ylim = c(y.min,y.max), xlab = '', ylab = '', axes = F, ty = 'n')


  #Add background
  u <- par('usr')
  rect(u[1], u[3], u[2], u[4], border = NA, col = bg.col)
  
  #Add end period shading
  if (endshade == TRUE){
    rect(end.start - 0.5, u[3], u[2], u[4], border = NA, col = end.col)
  }
  
  #Add mean line
  if (anomaly == F){
      if (mean_line == TRUE){
      abline(h = y.mean, col = 'grey', lwd = 3, lty = 2)
      } 
  } else if (anomaly == TRUE){
      abline(h = 0, col = 'grey', lwd = 3, lty = 2)
  }
  
  #Add x y lines
  abline(h = u[3], lwd=3)
  abline(v = u[1], lwd=3)
  
  #Add data points/lines
  points(x[, list(X, Value)], pch = 16, cex = point.cex)
  lines( x[, list(X, Value)], lwd = lwd)
  
  #extra lines
  if (extra == TRUE){
    x2 <- data[Var == y.var2, ]
    x2 <- x2[order(x2[, get(x.var2)]), ]
    setnames(x2, x.var2, 'X2')
    x2 <- x2[X2 >= x.start, ]
    if (mean_line.2 == TRUE){
     abline(h = mean(x2[, Value]), col = 'lightcoral', lwd = 3, lty = 2) 
    }
    points(x2[, list(X2, Value)], pch = 16, cex = point.cex, col = "indianred")
    lines( x2[, list(X2, Value)], lwd = lwd, col = "indianred")
    }
    
  
  #Add axis
  if (suppressAxis == FALSE){
    if(is.na(stacked)) axis(1, cex.axis = 1)
    if(!is.na(stacked)){
      if(stacked!= 'A') axis(3, cex.axis = 1.5, tck = 0.1, labels = F)
    }
  }

  #Stacked axes with 0 overlap so need to remove
  labels <- round((axTicks(2) / scale.axis), 5)
  if(labels[1] == 0) labels[1] <- ''
  axis(2, at = axTicks(2), labels = labels, cex.axis = rel.y.num,
       las = T)

    #Add axis labels
    if(!website){
      if(!is.na(stacked)) text(u[1], u[4], labels = stacked, cex = cex.stacked, adj = c(-0.5, 1.5))
    } else if (website){
      text(u[1], u[4], labels = "", cex = cex.stacked, adj = c(-0.5, 1.5))
    }
    if(is.na(stacked)){
      mtext(1, text = x.label, line = x.line, cex = 1)
      mtext(2, text = y.label, line = y.line, cex = rel.y.text)
    }
  
    if (full.trend == T){
    #Split data into past decade and full time series
    dat <- as.data.frame(x[, list(X, Value)])
    
    dat <- dat %>% dplyr::rename(series = Value) %>%
      mutate(time = seq(1,nrow(dat),1))
    
    # Fit linear model
    lm_out <- fit_lm(dat = dat)
    p <- lm_out$p
    if (p < .05){
        
      newtime <- seq(min(dat$time), max(dat$time), length.out=length(dat$time))
      newdata <- data.frame(time = newtime,
                      time2 = newtime^2)
      lm_pred <- AICcmodavg::predictSE(lm_out$model, 
                                 newdata = newdata,
                                 se.fit = TRUE)

      year <- seq(x$X[1],x$X[length(x$X)],length.out = length(dat$time))

      # Make plot
      if (lm_pred$fit[length(lm_pred$fit)] > lm_pred$fit[1]){
        lines(year, lm_pred$fit, col = main.pos, lwd = 7)
        points(x[, list(X, Value)], pch = 16, cex = point.cex)
        lines( x[, list(X, Value)], lwd = lwd)

        if (line.forward == TRUE){
           lines(year, lm_pred$fit, col = main.pos, lwd = 7)
        }
      } else if (lm_pred$fit[length(lm_pred$fit)] < lm_pred$fit[1]){
        lines(year, lm_pred$fit, col = main.neg, lwd = 7)
        points(x[, list(X, Value)], pch = 16, cex = point.cex)
        lines( x[, list(X, Value)], lwd = lwd)
        if (line.forward == TRUE){
           lines(year, lm_pred$fit, col = main.neg, lwd = 7)
        }
      }
    }
    
    if (extra == TRUE){
      
      # Second variable
      dat <- as.data.frame(x2[, list(X2, Value)])
    
      dat <- dat %>% dplyr::rename(series = Value) %>%
      mutate(time = seq(1,nrow(dat),1))
    
     # Fit linear model
      lm_out <- fit_lm(dat = dat)
      p <- lm_out$p
      points(x2[, list(X2, Value)], pch = 16, cex = point.cex, col = "indianred")
      lines( x2[, list(X2, Value)], lwd = lwd, col = "indianred")
      if (p < .05){
    
        newtime <- seq(min(dat$time), max(dat$time), length.out=length(dat$time))
        newdata <- data.frame(time = newtime,
                      time2 = newtime^2)
        lm_pred <- AICcmodavg::predictSE(lm_out$model, 
                                 newdata = newdata,
                                 se.fit = TRUE)

        year <- seq(x2$X2[1],x2$X2[length(x2$X2)],length.out =length(dat$time))
   
    # Make plot
        if (lm_pred$fit[length(lm_pred$fit)] > lm_pred$fit[1] ){
          lines(year, lm_pred$fit, col = main.pos, lwd = 7)
          points(x2[, list(X2, Value)], pch = 16, cex = point.cex, col = "indianred")
          lines( x2[, list(X2, Value)], lwd = lwd, col = "indianred")
        } else if (lm_pred$fit[length(lm_pred$fit)] < lm_pred$fit[1]){
          lines(year, lm_pred$fit, col = main.neg, lwd = 7)
          points(x2[, list(X2, Value)], pch = 16, cex = point.cex, col = "indianred")
          lines( x2[, list(X2, Value)], lwd = lwd, col = "indianred")
        } 
     }
    }

  }


 
}  


#Add axis labels for stacked plots
soe.stacked.axis <- function(x.label, y.label, x.line = 2.5,rel.x.text = 1.5,
                             y.line = 3.5, rel.y.text = 1.5, outer = TRUE){
  axis(1, cex.axis = rel.x.text)
  mtext(1, text = x.label, line = x.line, cex = rel.x.text, outer = outer)
  mtext(2, text = y.label, line = y.line, cex = rel.y.text, outer = outer)
  
}


#Background colors
background   <- 'white'
recent       <- '#E6E6E6'
main.pos <- rgb(253/255, 184/255, 99/255,  alpha = .9)
main.neg <- rgb(178/255, 171/255, 210/255, alpha = .9)

```

**Example plot**
```{r examplefig, fig.cap="Example plot formatted for the report", echo = T, eval = T}
opar <- par(mar = c(4, 6, 2, 6))

soe.plot(SOE.data.2018, "Time", "Recreational Seafood MA", scale.axis = 10^6,
         end.start = 2008, x.label = 'Year', rel.y.text = 1.5, rel.y.num = 1.1,
         y.line = 2.5, y.label = expression("Fish caught, 10"^6 *" n"))
```


