# MAFMC ABC/ACL and catch

**Description**: The catch limit (either ABC or ACL) and total catch from 2012 - 2020 for all MAFMC species and sector (commercial or recreational), if appropriate.

**Found in**:  State of the Ecosystem - Mid-Atlantic (2022) 

**Indicator category**: Synthesis of published information, Database pull

**Contributor(s)**: Jessica Coakley, Kiley Dancy, Jose Montanez, Julia Beaty, Karson Coutre, Jason Didden
  
**Data steward**: Brandon Muffley <bmuffley@mafmc.org>
  
**Point of contact**: Brandon Muffley <bmuffley@mafmc.org>
  
**Public availability statement**: Source data are publicly available

## Methods

### Data Sources

MAFMC Fishery Information Documents, Stock Assessment reports, SSC reports, GARFO catch/landings database, MRIP queries 

### Data Analysis

ABCs are set by the SSC, ACLs (if appropriate) developed by the Council; recreational data from MRIP, commercial catch from either the NEFSC assessment lead or GARFO database. Each species, depending upon data availability, sectors, fleets etc., goes through a different data processing process

Each species, depending upon data availability, sectors, fleets etc., goes through a different data analysis process.

### Data Processing 

Data were formatted for inclusion in the `ecodata` R package using the R code found [here](https://github.com/NOAA-EDAB/ecodata/blob/master/data-raw/get_abc.acl.R).

### Plotting

The plot below was built using the code found
[here](https://github.com/NOAA-EDAB/ecodata/blob/master/chunk-scripts/human_dimensions_MAB.Rmd-abcacl.R). 

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-abcacl.R"), eval = T, echo = F, fig.cap = "."}

```


```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-abcacl-catch.R"), eval = T, echo = F, fig.cap = "."}

```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-abcacl-stacked.R"), eval = T, echo = F, fig.cap = "."}

```
