# Aggregate Groups {#aggroups}

**Description**: Mappings of species into aggregate group categories for different analyses

**Found in**: State of the Ecosystem - Gulf of Maine & Georges Bank (2018+), State of the Ecosystem - Mid-Atlantic (2018+) 

**Indicator category**: Synthesis of published information 

**Contributor(s)**: Geret DePiper, Sarah Gaichas, Sean Hardison, Sean Lucey
  
**Data steward**: Sean Lucey <Sean.Lucey@noaa.gov>
  
**Point of contact**: Sean Lucey <Sean.Lucey@noaa.gov>
  
**Public availability statement**: Source data is available to the public (see Data Sources). 
```{r global-opts2, echo = FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## Methods
The State of the Ecosystem (SOE) reports are delivered to the New England Fishery Management Council (NEFMC) and Mid-Atlantic Fishery Management Council (MAFMC) to provide ecosystems context.  To better understand that broader ecosystem context, many of the indicators are reported at an aggregate level rather than at a single species level.  Species were assigned to an aggregate group following the classification scheme of @garrison2000dietary and @link2006EMAX.  Both works classified species into feeding guilds based on food habits data collected at the Northeast Fisheries Science Center (NEFSC).  In 2017, the SOE used seven specific feeding guilds (plus an "other" category; Table \@ref(tab:soe2017class)).  These seven were the same guilds used in @garrison2000dietary, which also distinguished ontogentic shifts in species diets.  

For the purposes of the SOE, species were only assigned to one category based on the most prevalent size available to commercial fisheries.  However, several of those categories were confusing to the management councils, so in 2018 those categories were simplified to five (plus "other"; Table \@ref(tab:soe2018class)) along the lines of @link2006EMAX.  In addition to feeding guilds, species managed by the councils have been identified.  This is done to show the breadth of what a given council is responsible for within the broader ecosystem context.

In the 2020 report, squids were moved from planktivores to piscivores based on the majority of their diet being either fish or other squid.  

```{r soe2017class, eval = T, echo = F}
soe.17.class <- data.frame('Feeding Guild' = c('Apex Predator', 'Piscivore', 
                                               'Macrozoo-piscivore', 'Macroplanktivore', 
                                               'Mesoplanktivore', 'Benthivore',
                                               'Benthos', 'Other'),
                           Description = c('Top of the food chain', 'Fish eaters', 
                                           'Shrimp and small fish eaters', 'Amphipod and shrimp eaters',
                                           'Zooplankton eaters', 'Bottom eaters',
                                           'Things that live on the bottom', 
                                           'Things not classified above'))

kable(soe.17.class, booktabs = TRUE,
      caption = "Aggregate groups use in 2017 SOE.  Classifications are based on Garrison and Link (2000). \\label{}")
```

```{r soe2018class, eval = T, echo = F}
soe.18.class <- data.frame('Feeding Guild' = c('Apex Predator', 'Piscivore', 
                                               'Planktivore', 'Benthivore',
                                               'Benthos', 'Other'),
                           Description = c('Top of the food chain', 'Fish eaters', 
                                           'Zooplankton eaters', 'Bottom eaters',
                                           'Things that live on the bottom', 
                                           'Things not classified above'))

kable(soe.18.class, booktabs = TRUE,
      caption = "Aggregate groups use since 2018 SOE.  Classifications are based on Link et al. (2006).")
```

### Data sources

In order to match aggregate groups with various data sources, a look-up table was generated which includes species' common names (COMNAME) along with their scientific names (SCINAME) and several species codes. SVSPP codes are used by the NEFSC Ecosystems Surveys Branch (ESB) in their fishery-independent Survey Database (SVDBS), while NESPP3 codes refer to the codes used by the Commercial Fisheries Database System (CFDBS) for fishery-dependent data. A third species code provided is the ITISSPP, which refers to species identifiers used by the Integrated Taxonomic Information System (ITIS). Digits within ITIS codes are hierarchical, with different positions in the identifier referring to higher or lower taxonomic levels. More information about the SVDBS, CFDBS, and ITIS species codes are available in the links provided below.

Management responsibilities for different species are listed under the column "Fed.managed" (NEFMC, MAFMC, or JOINT for jointly managed species). More information about these species is available on the FMC websites listed below. Species groupings listed in the "NEIEA" column were developed for presentation on the Northeast Integrated Ecosystem Assessment (NE-IEA) website. These groupings are based on EMAX groupings [@link2006EMAX], but were adjusted based on conceptual models developed for the NE-IEA program that highlight focal components in the Northeast Large Marine Ecosystem (i.e. those components with the largest potential for perturbing ecosystem dynamics). NE-IEA groupings were further simplified to allow for effective communication through the NE-IEA website.

#### Supplemental information

See the following links for more information regarding the NEFSC ESB Bottom Trawl Survey, CFDBS, and ITIS:  

*    https://www.itis.gov/  
*    https://inport.nmfs.noaa.gov/inport/item/22561  
*    https://inport.nmfs.noaa.gov/inport/item/22560  
*    https://inport.nmfs.noaa.gov/inport/item/27401  	

More information about the NE-IEA program is available [here](http://integratedecosystemassessment.noaa.gov).

More information about the New Engalnd Fisheries Management Council is available [here](https://www.nefmc.org/).

More information about the Mid-Atlantic Fisheries Management Council is available [here](http://www.mafmc.org/).

### Data extraction 

Species lists are pulled from SVDBS and CFDBS.  They are merged using the ITIS code.  Classifications from Garrison and Link [@garrison2000dietary] and Link et al. [@link2006EMAX] are added manually. The R code used in the extraction process can be found [here](https://github.com/slucey/RSurvey/blob/master/Species_list.R).






