# Expected Number of Species

**Description**: Time Series of Expected Number of Species per Tow in NEFSC BTS

**Found in**: State of the Ecosystem - Gulf of Maine & Georges Bank (2021+), State of the Ecosystem - Mid-Atlantic (2021+)

**Indicator category**: Database pull with analysis; Published methods

**Contributor(s)**: Sean Lucey
  
**Data steward**: Sean Lucey, <sean.lucey@noaa.gov>
  
**Point of contact**: Sean Lucey, <sean.lucey@noaa.gov>
  
**Public availability statement**: 


  
## Methods


### Data sources



### Data analysis


### Data processing
Data were formatted for inclusion in the `ecodata` R package using the R code found [here](https://github.com/NOAA-EDAB/ecodata/blob/master/data-raw/get_exp_n.R).

### Plotting

The plot below was built using the code found
[here](https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-exp-n.R). 

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-exp-n.R"), eval = T, echo = F, fig.cap = "Expected number of species per 1000 individuals for FAll NEFSC bottom trawl survey."}

```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-exp-n-spring.R"), eval = T, echo = F, fig.cap = "Expected number of species per 1000 individuals for Spring NEFSC bottom trawl survey."}

```

