# Highly Migratory Species Landings

**Description**: Highly Migratory Species Landings

**Found in**: State of the Ecosystem - Gulf of Maine & Georges Bank (2020), State of the Ecosystem - Mid-Atlantic (2020)

**Indicator category**: Synthesis of published information, Database pull with analysis

**Contributor(s)**: George Silva, Heather Baertlein, and Cliff Hutt
  
**Data steward**: Kimberly Bastille
  
**Point of contact**: George Silva <george.silva@noaa.gov>
  
**Public availability statement**: Source data are publicly available.

## Methods


### Data sources
Data from eDealer database and Bluefin Tuna Dealer reports on SAFIS.


### Data extraction 
Data was processed for Fisheries of the United States and then lumped by region to avoid confidentiality issues.

### Data analysis
Data was analyzed using Excel pivot tables.

### Data processing
HMS landings data were formatted for inclusion in the `ecodata` R package using the following R code.
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/data-raw/get_hms_landings_rev.R"), eval = F, echo = T}

```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/data-raw/get_hms_landings_weight.R"), eval = F, echo = T}

```
### Plotting

