# Harbor Porpoise and Gray Seal Bycatch


**Description**: Harbor Porpoise and Gray Seal Indicator

**Found in**: State of the Ecosystem - Gulf of Maine & Georges Bank (2018+), State of the Ecosystem - Mid-Atlantic (2018+)

**Indicator category**: Synthesis of published information; Published methods

**Contributor(s)**: Christopher D. Orphandies, Debra Palka
  
**Data steward**: Debra Palka <debra.palka@noaa.gov>

**Point of contact**: Debra Palka <debra.palka@noaa.gov>

**Public availability statement**: Source data are available in public [stock assessment reports](https://www.fisheries.noaa.gov/national/marine-mammal-protection/marine-mammal-stock-assessment-reports-region).
  

## Methods


### Data sources
Reported harbor porpoise bycatch estimates and potential biological removal levels can be found in publicly available documents; detailed in [Marine Mammal Protection Stock Assessments](https://www.fisheries.noaa.gov/national/marine-mammal-protection/marine-mammal-stock-assessment-reports-region). More detailed documentation as to the methods employed can be found in NOAA Fisheries Northeast Fisheries Science Center (NEFSC) Center Reference Documents (CRDs) found on the NEFSC [publications page](https://www.nefsc.noaa.gov/publications/crd). 


### Data extraction 
Annual gillnet bycatch estimates are documented in the CRDs. These feed into the Stock Assessment Reports which report both the annual bycatch estimate and the mean 5-year estimate. 

### Data analysis
Bycatch estimates as found in stock assessment reports were plotted along with confidence intervals. The confidence intervals were calculated from published CVs assuming a normal distribution ($\sigma = \mu CV$; $CI = \bar{x} \pm \sigma * 1.96$).

Data were analyzed and formatted for inclusion in the `ecodata` R package using the R code found here, [Harbor Porpoise](https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/data-raw/get_harborporpoise.R) and [Gray Seal](https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/data-raw/get_grayseal.R).


### Plotting
Code used to plot data can be found here, [Harbor Porpoise](https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-harborporpoise.R) and [Gray Seal](https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-grayseal.R).

```{r , code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-harborporpoise.R"), eval = T, echo=F, fig.cap="Estimated Harbor porpoise bycatch and the potential biological removal."}

```

```{r , code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-grayseal.R"), eval = T, echo=F, fig.cap="Estimated Gray Seal bycatch and the potential biological removal."}

```
