# Marine Heatwave

**Description**: Marine Heatwave

**Found in**: State of the Ecosystem - Gulf of Maine & Georges Bank (2020+), Mid-Atlantic (2020+)

**Indicator category**: Published methods, Database pull with analysis

**Contributor(s)**: Vincent Saba
  
**Data steward**: Kimberly Bastille <kimberly.bastille@noaa.gov>
  
**Point of contact**: Vincent Saba <vincent.saba@noaa.gov>
  
**Public availability statement**: 

## Methods
Marine heatwave analysis for Georges Bank, Gulf of Maine, and the Middle Atlantic Bight according to the definition in @hobday2016.  Heatwaves are defined as temperatures that exceed the 90th percentile for at least 5 consecutive days for surface heatwaves and at least 30 consecutive days for bottom heatwaves. 

### Data sources
[NOAA high-res OISST (daily, 25-km, 1982-2022)](https://www.esrl.noaa.gov/psd)

### Data extraction 
Each yearly file (global) was downloaded, concatenated into a single netcdf file using nco (Unix), and then cropped to the USNES region using Ferret.  Each EPU's time-series of SST was averaged using .shp file boundaries for the MAB, GB, and GOM (also done in Ferret) and saved to the three .csv files.

### Data analysis

Changes in the marine heatwave analysis in 2023 are outlined below. 

#### Detrended Data

Before analysis was conducted, the data was detrended in order to capture extreme events by removing the overall warming temperature trend.


#### Climatology

In 2023 the climatological period shifted from 1982-2011 to 1982-2022, now using the whole time series as the climatology. 

#### Indictaors

In reports prior to 2023, heatwave Maximum Intensity (maximum temperature observed during a
heatwave event - deg. C) and Cumulative Intensity (sum of daily intensity anomalies - deg. C x days) were include. We now include Maximum Intensity and Number of Days in a heatwave state (N days).


The marine heatwave metrics  are calculated using NOAA OISST daily sea surface temperature data (25-km resolution) from January 1982 to December 2022.  The heatwaves are calculated based on the algorithms in Hobday et al. 2016 and by using a climatology of 1982-2022.  These metrics were run using [`heatwaveR`](https://robwschlegel.github.io/heatwaveR/).

### Data processing 
Marine Heatwave data were formatted for inclusion in the `ecodata` R package using this [R code](https://github.com/NOAA-EDAB/ecodata/blob/master/data-raw/get_heatwave.R).

### Plotting

Code used for the plots below can be found [here](https://github.com/NOAA-EDAB/ecodata/blob/master/chunk-scripts/LTL_MAB.Rmd-heatwave.R).


```{r , code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-heatwave.R"), eval = T, echo=F, fig.cap="Cumulative and maximum marine heatwave in the Mid-Atlantic", fig.align="center"}

```


```{r, fig.cap="Mid-Atlantic Marine Heatwave events in 2021.",code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-heatwave-year.R"), eval = T, echo=F}

```
