# Ocean Acidification

**Description**: Ocean Acidification 

**Indicator category**: 

**Found in:** State of the Ecosystem - Gulf of Maine & Georges Bank (2021); State of the Ecosystem - Mid-Atlantic Bight (2021)

**Contributor(s)**: Grace Saba 

**Data steward**: Grace Saba

**Point of contact**: Grace Saba <saba@marine.rutgers.edu>

**Public availability statement**: 

## Methods

The work included in the State of the Ecosystem came directly from @Wright2020. 

### Data sources

See @Wright2020.

### Data extraction

See @Wright2020.


### Data analysis

See @Wright2020.


### Data processing

See @Wright2020.


### Plotting

Code for plotting can be found [here](https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-mab-oa-glider.R). 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-mab-oa.R"), eval=TRUE, echo = FALSE}

```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-mab-oa-glider.R"), eval=TRUE, echo = FALSE}

```


