# Right Whale Abundance 


```{r,  echo = F, message=F}

#Load packages
library(knitr)
library(rmarkdown)

```
**Description**: Right Whale

**Found in**: State of the Ecosystem - Gulf of Maine & Georges Bank (2017, 2018, 2019), State of the Ecosystem - Mid-Atlantic (2017, 2018, 2019)

**Indicator category**: Synthesis of published information; Published methods

**Contributor(s)**: Christopher D. Orphanides
  
**Data steward**: Chris Orphanides, chris.orphanides@noaa.gov
  
**Point of contact**: Richard Pace, richard.pace@noaa.gov
  
**Public availability statement**: Source data are available from the New England Aquarium upon request. Derived data are available [here](http://comet.nefsc.noaa.gov/erddap/tabledap/protected_species_soe_v1.html)

## Methods

### Data sources
The North Atlantic right whale abundance estimates were taken from a published document [see @Pace2017], except for the most recent 2016 and 2017 estimates. Abundance estimates from 2016 and 2017 were taken from the 2016 NOAA marine mammal stock assessment [@Hayes2017] and an unpublished 2017 stock assessment.


### Data extraction 
Data were collected from existing reports and validated by report authors. 

### Data analysis
Analysis for right whale abundance estimates is provided by @Pace2017, and code can be found in the [supplemental materials](https://onlinelibrary.wiley.com/action/downloadSupplement?doi=10.1002%2Fece3.3406&file=ece33406-sup-0001-SupInfo.docx). 

### Plotting

```{r rw-abundance, fig.cap="North Atlantic right whale population estimates shown with 95% credible intervals.",echo = T, message=F, warning=F,  fig.pos='H', fig.height = 4.5, fig.width=7}

# Relative working directories
data.dir  <- here::here('data')
r.dir <- here::here('R')

# Load data
load(file.path(data.dir,"SOE_data_erddap.Rdata"))

# Source plotting functions
source(file.path(r.dir,"BasePlot_source.R"))

par(mar = c(5,5,4,1))
soe.plot(SOE.data, 'Time', "right_whale_median", rel.y.num = 1.2, 
         full.trend = F, lwd = 1.5, point.cex = 1, end.start = 2007,
         x.label = 'Year', y.label = 'Abundance, n', 
         rel.y.text = 1)

lw_CI <- SOE.data[Var == 'right_whale_lower_95', list(Time, Value)]
up_CI <- SOE.data[Var == 'right_whale_upper_95', list(Time, Value)]
points(lw_CI, type  = "l", lty = 2, col = adjustcolor("darkorange", .6), lwd = 2.5)
points(up_CI, type  = "l", lty = 2, col = adjustcolor("darkorange", .6), lwd = 2.5)

```