# Right Whale Abundance 


```{r,  echo = F, message=F}

#Load packages
library(knitr)
library(rmarkdown)

```
**Description**: Right Whale

**Found in**: State of the Ecosystem - Gulf of Maine & Georges Bank (2017, 2018, 2019, 2020), State of the Ecosystem - Mid-Atlantic (2017, 2018, 2019, 2020)

**Indicator category**: Synthesis of published information; Published methods

**Contributor(s)**: Christopher D. Orphanides
  
**Data steward**: Chris Orphanides, chris.orphanides@noaa.gov
  
**Point of contact**: Richard Pace, richard.pace@noaa.gov
  
**Public availability statement**: Source data are available from the New England Aquarium upon request. Derived data are available [here](http://comet.nefsc.noaa.gov/erddap/tabledap/protected_species_soe_v1.html)

## Methods

### Data sources
The North Atlantic right whale abundance estimates were taken from a published document [see @Pace2017], except for the most recent 2016 and 2017 estimates. Abundance estimates from 2016 and 2017 were taken from the 2016 National Oceanographic and Atmospheric Administration marine mammal stock assessment [@Hayes2017] and an unpublished 2017 stock assessment.

Calves birth estimates are taken from a published report [@narw2019] put out yearly by the North American Right Whale Consortium. 

### Data extraction 
Data were collected from existing reports and validated by report authors. 

### Data analysis
Analysis for right whale abundance estimates is provided by @Pace2017, and code can be found in the [supplemental materials](https://onlinelibrary.wiley.com/action/downloadSupplement?doi=10.1002%2Fece3.3406&file=ece33406-sup-0001-SupInfo.docx). 

### Data processing

Time series of right whale  and calf abundance estimates were formatted for inclusion in the `ecodata` R package using this R [code](https://github.com/NOAA-EDAB/ecodata/blob/master/data-raw/get_narw.R).



### Plotting
Code used create the plots below can be found at these links, NARW [population estimates](https://github.com/NOAA-EDAB/ecodata/blob/master/chunk-scripts/macrofauna.Rmd-NARW-abundance.R) and [calf births](https://github.com/NOAA-EDAB/ecodata/blob/master/chunk-scripts/macrofauna.Rmd-NARW-calf-abundance.R).


```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna.Rmd-NARW-abundance.R"), fig.cap="North Atlantic right whale population estimates shown with 95% credible intervals.",echo = F, message=F, warning=F,  fig.pos='H', fig.height = 4.5, fig.width=7}

```


```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna.Rmd-NARW-calf-abundance.R"), fig.cap="North Atlantic right whale calf births.",echo = F, message=F, warning=F,  fig.pos='H', fig.height = 4.5, fig.width=7}

```

