# Single Species Status Indicator {#stockstatus}


**Description**: Summary of the most recent stock assessment results for each assessed species.

**Found in**: State of the Ecosystem - Gulf of Maine & Georges Bank (2017, 2018, 2019, 2020), State of the Ecosystem - Mid-Atlantic (2017, 2018, 2019, 2020)

**Indicator category**: Synthesis of published information

**Contributor(s)**: Sarah Gaichas, based on code and spreadsheets originally provided by Chris Legault

**Data steward**: Sarah Gaichas <sarah.gaichas@noaa.gov>

**Point of contact**: Sarah Gaichas <sarah.gaichas@noaa.gov>

**Public availability statement**: All stock assessment results are publicly available (see Data Sources). Summarized data are available [here](http://comet.nefsc.noaa.gov/erddap/tabledap/assess_soe_v1.htmlTable?No,Entity_Name,Science_Center,Assessment_Year,Last_Data_Year,Assessment_Level,Citation,Comments,Best_F,F_Year,Flimit,Fmsy,F_Flimit,F_Fmsy,Best_B,B_Year,B_Blimit,B_Bmsy,Stock_Level_Relative_to_Bmsy,Bmsy,Blim).

## Methods

### Data sources
<!--Please provide a text description of data sources, inlcuding primary collection methods. What equipment was used to turn signal to data? From which vessel were data collected and how? What quality control procedures were employed, if any?--> 

"Data" used for this indicator are the outputs of stock assessment models and review processes, including reference points (proxies for fishing mortality limits and stock biomass targets and limits), and the current fishing mortality rate and biomass of each stock. The spreadsheet summarizes the most recent stock assessment updates for each species, which are available on the Northeast Fisheries Science Center (NEFSC) website at:
https://www.nefsc.noaa.gov/saw/reports.html  
https://www.nefsc.noaa.gov/publications/crd/crd1717/

Additional assessments are reported directly to the New England Fishery Management Council (NEFMC):
http://s3.amazonaws.com/nefmc.org/Document-2-SAFE-Report-for-Fishing-Year-2016.pdf  
http://s3.amazonaws.com/nefmc.org/4_NEFSC_SkateMemo_July_2017.pdf



### Data extraction

Each assessment document was searched to find the following information (often but not always summarized under a term of reference to determine stock status in the executive summary):

*    **Bcur**: current year biomass, (most often spawning stock biomass (SSB) or whatever units the reference points are in)

*    **Fcur**: current year fishing mortality, F

*    **Bref**: biomass reference point, a proxy of Bmsy (the target)

*    **Fref**: fishing mortality reference point, a proxy of Fmsy

### Data processing

R code used to process the stock status data set for inclusion in the `ecodata` R package can be found [here](https://github.com/NOAA-EDAB/ecodata/blob/master/data-raw/get_stocks.R).

### Data analysis
<!--Text description of analysis methods, similar in structure and detail to a peer-reviewed paper methods section-->

For each assessed species, Bcur is divided by Bref and Fcur is divided by Fref. They are then plotted for each species on an x-y plot, with Bcur/Bref on the x axis, and Fcur/Fref on the y axis. 

<!--What packages or libraries did you use in your work flow?-->
```{r, eval=F, echo = F}
#all packages in the SOE file that ran my code snippet, probably only need data.table
library(Kendall);library(data.table);library(zoo)
library(dplyr);library(nlme);library(AICcmodavg)
library(colorRamps);library(Hmisc);library(rgdal)
library(maps);library(mapdata);library(raster)
library(grid);library(stringr);library(png)
library(ncdf4);library(marmap); library(magick)


```

<!--Include accompanying R code, pseudocode, flow of scripts, and/or link to location of code used in analyses.-->

### Plotting

The script used to develop the figure in the State of the Ecosystem report is below, with an example figure. Different lines are commented out of the script to produce the Mid- Atlantic or New England figures. The positioning table used to make the plot (the data frame named "decoder" below) is available [here](http://comet.nefsc.noaa.gov/erddap/tabledap/assess_support_soe_v1.htmlTable?Entity_Name,Count_of_No,Min_of_No2,Code,Council,my_pos,my_pos2,my_pos3).

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions.Rmd-stock-status1.R"), warning=F, fig.cap ="Summary of single species status for MAFMC and jointly managed stocks.", fig.width = 8, fig.asp = 0.65, eval=T, echo=F}

```

