# Fisheries Revenue in Wind Development Areas {#wind_revenue}

**Description**: Top Species Revenue from Wind Development Areas

**Found in**: State of the Ecosystem - Gulf of Maine & Georges Bank 2022+ (Different Methods 2021), State of the Ecosystem - Mid-Atlantic 2022+ (Diferent Methods 2021)

**Indicator category**: Database pull with analysis

**Contributor(s)**: Benjamin Galuardi, Douglas Christel
  
**Data steward**: Doug Christel <douglas.christel@noaa.gov>
  
**Point of contact**: Doug Christel <douglas.christel@noaa.gov>
  
**Public availability statement**: Source data are NOT publicly available. Please email douglas.christel@noaa.gov for further information and queries of indicator source data.

## Methods

### Data Sources

Modeled vessel trip report (VTR) data using the fishing footprint method (DePiper 2014 and Benjamin et al. 2017) linked with dealer reports for annual landings and revenue within wind lease areas and dealer report data for annual GARFO landings/revenue.


### Data Analysis

Using raster data of modeled VTR data using the Fishing Footprint method, we integrated dealer data and compared to existing/proposed wind lease areas to get landings/revenue in each area by year.  

### Data Processing 

Data were formatted for inclusion in the `ecodata` R package using the R code found [here](https://github.com/NOAA-EDAB/ecodata/blob/master/data-raw/get_wind_revenue.R).

### Plotting

The plot below was built using the code found
[here](https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-wea-spp-rev.R). 

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-wea-spp-rev.R"), eval = T, echo = F, fig.cap = "Top five MAFMC species revenue found in wind development areas."}

```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-wea-landings-rev.R"), eval = T, echo = F, fig.cap = "Species landings revenue range found in wind development areas."}

```

\* Landings and revenue for these species are likely underestimated due to limited coverage of these fisheries in historic reporting requirements for vessels issued federal permits by the NMFS Greater Atlantic Regional Fisheries Office.  However, such limitations also suggest an inaccurately higher proportion of such landings and revenues in existing lease areas. ** Clearnose skates were reported separately from skates, which is presumed to include all skates managed under the Northeast skate complex. *** Based on comparison with other data sources, the high values for Illex squid are likely overestimates affected by the methods used to model logbook data to estimate spatial overlap of fishing operations with wind energy areas.


<!-- ```{r,  eval = T, echo = F, fig.cap = "Species landings revenue range found in wind development areas."} -->
<!-- dt<- ecodata::wea_landings_rev %>%  -->
<!--   dplyr::select("NEFMC, MAFMC, and ASMFC Managed Species",  -->
<!--                 "perc_landings_max","perc_revenue_max" ) %>%  -->
<!--  dplyr::slice_head(n =10) %>%  -->
<!--   dplyr::mutate(perc_landings_max = paste0(perc_landings_max, " %"), -->
<!--                 perc_revenue_max = paste0(perc_revenue_max, " %")) %>%  -->
<!--   dplyr::rename("Maximum Percent Total Annual Regional Species Landings"="perc_landings_max", -->
<!--     "Maximum Percent Total Annual Regional Species Revenue"="perc_revenue_max")  -->
<!-- k<- kable(dt, caption = "Top ten species Landings and Revenue from Wind Energy Areas." ) %>%  -->
<!--   kable_classic(full_width = F, html_font = "Cambria") -->
<!-- kableExtra::landscape(k) -->
<!--``` -->



## Methods 2021
### Data Sources

This indicator is derived from the data underpinning the "Sociceoeconomic Impacts of Atlantic Offshore Wind Development" web site, which can be accessed at https://www.fisheries.noaa.gov/resource/data/socioeconomic-impacts-atlantic-offshore-wind-development.

The underlying raster data is defined in Benjamin S, Lee MY, DePiper G. 2018. Visualizing fishing data as rasters. NEFSC Ref Doc 18-12; 24 p.

This raster data was then linked to the Greater Atlantic Regional Office's Data Matching Imputation System (https://www.fisheries.noaa.gov/inport/item/17328) to derive revenue estimates from the Wind Energy Areas, defined as of December 11, 2020. Of note is that the version of DMIS utilized for this reporting includes the SFCLAM data missing from the traditional DMIS dataset. All revenue estimates are deflated to 2019 dollars using the St. Louis Federal Reserve's Quarterly Implicit GDP Deflator, which can be accessed at https://fred.stlouisfed.org/data/GDPDEF.txt

### Data Analysis
Code used to analyze this data can be [found here](https://github.com/NOAA-EDAB/tech-doc/blob/master/R/stored_scripts/WindRevenue_Code_for_Dissemination.R)



