# Wind Energy Development Timeline

**Description**: Wind Energy Lease Area Development 

**Found in**: State of the Ecosystem - Gulf of Maine & Georges Bank (2021+), State of the Ecosystem - Mid-Atlantic (2021+)

**Indicator category**: Published methods, Synthesis of published information, Database pull, Database pull with analysis

**Contributor(s)**: Angela Silva, Andrew Lipsky, Doug Christel
  
**Data steward**: Angela Silva
  
**Point of contact**: Angela Silva <angela.silva@noaa.gov>
  
**Public availability statement**: Source data are NOT publicly available. Please email angela.silva@noaa.gov for further information and queries of Speed and Extent of Offshore Wind Development indicator source data.

## Methods
### Data Sources
BOEM lease area, Call Areas, Planning Area shapefiles:  https://www.boem.gov/renewable-energy/mapping-and-data/renewable-energy-gis-data; 

Maine Area of Interest: Maine Department of Marine Resources, Central Atlantic Bight planning area draft (BOEM communication, INTERNAL ONLY private shapefile); Foundation and Cable data from South Fork Final Environemntal Impact Statement (SWFW FEIS) data tables E-4, E-4-1, E-2: https://www.boem.gov/sites/default/files/documents/renewable-energy/state-activities/SFWF%20FEIS.pdf


### Data Analysis
All data was updated for 2022 with South Fork Wind Farm FEIS and the following assumptions were made on future wind areas: 
* (1) There are no reported values for foundations, cable acres and miles and year of construction for NY WEA, Maine AOI, and Central Atlantic Bight draft planning area. 
* (2) To estimate the variables, the ratio of each (Cumul_FNDS, Cumul_Offsh_Cbl_Acres, Cumul_OffExp_Inter_Cab_Miles, TBNSinstall_no) was calculated by using reported values for existing lease area. All data is reported as ""2030""

Spatial Analysis for Project_Acres:


Project Areas and Call Area acres were calculated using BOEM Project Area Shapefiles (Project_Areas_12_3_2019), BOEM NY Call Area Shapefiles (NY_Call_Areas), and NY Call Area Primary and Secondary Recommendation shapefiles (BOEM_NY_Draft_WEAs_11_1_2018) in ArcMap. 


Project_Areas_12_3_2019, NY_Call_Areas, and BOEM_NY_Draft_WEAs_11_1_2018 Acres were calculated using Add Field and Field Calculator tool. Python Expression = !shape.area@acres!


Project_Name:  Table E-4 of South Fork FEIS Project names were matched to shapefiles by name and lease number. 


FDNS: Number of foundations proposed or expected for each Project area taken directly from Table E-4 of South Fork DEIS.


Offsh_Cbl_Acres: Values taken directly from Table A-4 in South Fork DEIS (Table A-4: Offshore Wind Leasing Activities in the U.S. East Coast: Projects and Assumptions [part 2], pg. E-3-4). Total values for MA/RI lease areas Bay State Wind, Liberty Wind, OCS-A 0522 Remainder, OCS-A 0500 Remainder, OCS-A 0521 Remainder, OCS-A 0520 were aggregated in the table (567 total acres).  Values were evenly distributed across the 6 Project areas.  As such, these values should be treated as estimates until more information is released specific to individual project areas and their landing sites. 


Dominion Energy was presented as 3 phases in Table E-4 for Project_Name (Dominion Energy Phase1, Dominion Energy Phase 2, Dominion Energy Phase 3). Only one Project shapefile area exists for this lease  area OCS-A 0483. Therefore, the total shapefile acreage was evenly divided between 3 phases similar to how the foundations were treated in table E-4 (Future Offshore Wind Project Construction Schedule, pg. E-14). 


OffExpCab_Miles: Offshore Export Cable Length OCS-A 0482, OCS-A 0519 OCS-A 0490 had 360 offshore export cable miles reported in Table E-4. This number was divided by 3 and 120 were assigned to these three project areas. 


### Data Processing 

Data were formatted for inclusion in the `ecodata` R package using the R code found [here](https://github.com/NOAA-EDAB/ecodata/blob/master/data-raw/get_wind_dev_speed.R).

### Plotting

The plot below was built using the code found
[here](https://github.com/NOAA-EDAB/ecodata/blob/master/chunk-scripts/human_dimensions_MAB.Rmd-wind-dev-speed0.30.R).

#### Proposed Wind Development Timeline

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-wind-proposed-dev.R"), eval = T, echo = F, fig.cap = "Cumulative area of wind lease sites being developed on the Northeast Shelf."}

```

#### Wind Development Map

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-wind-dev-cumul.R")}

```

#### Wind and Scientific Surveys

<!-- ```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-wind-survey-table.R"), fig.cap="Survey mitigation timeline."} -->

<!-- ``` -->


```{r, fig.cap="Survey mitigation timeline."}
dat<- data.frame(XSurvey= c("Fall BTS", "Spring BTS", "EcoMon", "Scallop",
                            "Shellfish(Clams)", "Right Whale (Air)", 
                            "Marine Mammal/Turtle (Ship/Air)",
                            "Altantic Shark (Bottom Long-Line",
                            "GOM Bottom Long-Line", "GOM Shrimp Survey", 
                            "Atlantic Shark COASTPAN"), 
                 X1 = c("Started", "Started", "No",
                                                    "Started", "No", "Inital", 
                                                    "No", "No","No", "No","No"), 
                 X2 = c("Inital", "Initial", "No", "Initial", 
                                            "No", "Initial","No", "No","No", "No",
                                            "No" ), 
                 X3 = c( "No", "No","No",
                        "No","No",  "Initial", "No", "No","No", "No","No"), 
                 X4 = c("No", "No","No", "No","No","No", 
                                         "No","No", "No","No","No"), 
                 X5 = c("No", "No","No", "No","No","No", 
                                         "No","No", "No","No","No"),
                 X6 = c("Initial", "Initial","No", "No","No","No", 
                                         "No","No", "No","No","No"))

column_names <- c("Survey","1.Evaluate designs & Impacts","2.Design New Methods","3.Calibrate New/Existing Surveys","4.Bridge Solutions","5.Conduct New Surveys","6.Comms & Data")
#DT::datatable(dat, colnames = column_names)

kable(dat, col.names = column_names) %>% 
  kableExtra::column_spec(1:length(colnames(dat)), width = "2cm") %>% 
  landscape()

```





